\documentclass[12pt]{amsart}
\usepackage[left=4cm,top=3cm,right=3cm,left=3cm]{geometry} 
\geometry{a4paper}     
\usepackage{fullpage}
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{tipa}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{epstopdf}
\usepackage{setspace}
\usepackage{palatino, url, multicol}
\usepackage{hanging}

\newcommand{\word}[5]{\par\textbf{#4}: $_{#5}$ \textit{#3} #1 [\textipa{#2}]}
\newcommand{\lenite}[6]{\par\textbf{#4}: $_{#6}$ \textit{#3} #1 [\textipa{#2}] \textit{allomorph of} #5}
%\newcommand{\marker}[4]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3}
\newcommand{\cw}[9]{\par\textbf{#4}: $_{#9}$ \textit{#3} #1 [\textipa{#2}] \textit{(c.w. from} \textbf{#6} #5 \textit{and} \textbf{#8} #7)}
\newcommand{\cww}[7]{\par\textbf{#4}: $_{#7}$ \textit{#3} #1 [\textipa{#2}] \textit{(c.w. from} \textbf{#6} #5)}
\newcommand{\loan}[6]{\par\textbf{#4}: $_{#6}$ \textit{#3} #1 [\textipa{#2}] \textit{(loaned from Eng.} \textbf{#5})}
\newcommand{\note}[8]{\par\textbf{#4}: $_{#8}$ \textit{#3} #1 [\textipa{#2}] (\textit{#5} \textbf{#7} #6)}

\newcommand{\derive}[9]{\par\textbf{#4}: $_{#9}$ \textit{#3} #1 [\textipa{#2}] \textit{derived from} \textbf{#6} #5 \textit{and} \textbf{#8} #7}
\newcommand{\derives}[7]{\par\textbf{#4}: $_{#7}$ \textit{#3} #1 [\textipa{#2}] \textit{derived from} \textbf{#6} #5}

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{English - Na'vi Dictionary}
\author{Taronyu : Richard Littauer}
\date{Last updated: January 17$^{th}$, 2010}

\begin{document}
\newpage
\singlespacing
\maketitle
This is an English to Na'vi dictionary. Abuse it well. Eternal thanks be to Tuiq for alphabetising this.\\

\textbf{Bold} indicates Na'vi word. {\sl Italics} indicate English definition. \underline{Underlining} indicates where Frommer has given us stress information, which is not otherwise predictable: I have included this in the IPA transcriptions instead of using normal IPA \textipa{"}, because syllable boundaries are often hard to place - any fault in underlining, unless a typing error, can be laid at Frommer's door. Plain text indicates grammatical categorisation: parentheses indicate further notes. $<$x$>$ represents an infix, where {\it t}$<$1$>$$<$2$>${\it ar}$<$3$>${\it on} are the positions. Proper nouns have been capitalised. -- and + indicates that an affix or adposition can fall on either side. Knowledge of lenition processes will be assumed. Elision sometimes occurs, but only in song.For more on inflections, see \href{http://www.talknavi.com/taronyu/Inflections.pdf}{this document.}\\ \\
{\bf Abbreviations:}
\begin{multicols}{2}{\noindent
-- = morpheme boundary\\
+ = leniting morpheme boundary\\
= = eliding boundary\\
$< >$ = infix morpheme\\
` = ejective [in IPA]\\
adj. = adjective\\
adp. = adpositional affix\\
adv. = adverb\\
conj. = conjunction\\
c.w. = compound word\\
intj. = interjection\\
inter. = interrogative\\
lit. = literally\\
n. = noun\\
num. = number\\
part. = particle\\
prep. = preposition\\
pn. = pronoun\\
prop.n. = proper noun\\
v. = verb\\
$\phi$ = illegal phonotactics}
\end{multicols}

{\bf Sources}
\begin{itemize}
\item F = Frommer himself.
\item FE = Frommer's personal emails
\item MS = the movie script
\item S = the list at the end of the Survival Guide
\item W = Wikipedia (which has received Frommer's blessing)
\item SG = the Survival Guide main text (flora and songs)
\item M = transcriptions or scripts from the Avatar movie
\item C = Cameron himself (from the 2007 script)
\item T = Taronyu
\item D  = words not derived by Taronyu.
\item G = the Avatar game (which has illegal names)
\item SG = the Survival Guide main text (everything else)

\end{itemize}

\newpage

\noindent\textbf{A}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{tx\`ing}{t'IN}{v.}{abandon}{S}
\derive{*(')em}{(P)Em}{prep.}{above}{ta'em}{from above}{emza'u}{pass}{T}
\word{ka}{ka}{prep.}{across}{S}
\derive{*kem}{kEm}{}{action}{f\`ikem}{this action}{kempe}{what action}{T}
\word{n\`imun}{nImun}{adv.}{again}{S}
\word{mllte}{m\textsyllabic{l}:\underline{tE}}{v.}{agree}{S}
\word{rawke}{RawkE}{n.}{alarm cry}{S}
\word{faketuan}{fakEtuan}{n.}{alien}{?} (might be poor transcription of {\bf fa} + {\bf ketuan}) $_{T}$
\word{ketuwong}{\underline{kE}tuwong}{n.}{alien}{S}
\word{kewong}{\underline{kE}woN}{adj.}{alien}{S}
\word{n\`iwotx}{nIwot'}{adv.}{all (of)}{S}
\word{tung}{tuN}{v.}{allow}{S}
\word{stum}{stum}{adv.}{almost}{S}
\word{n\`i'awtu}{nIP\underline{aw}tu}{adv.}{alone}{S}
\word{kip}{kip}{prep.}{among}{S}
\derive{pizayu}{\underline{pi}zaju}{n.}{ancestor}{*za}{come}{yu}{noun deriving affix $_T$}{S}
\note{s\`i=}{sI}{conj.}{and}{connects two things: for clauses use}{ulte}{and}{S}
\word{ulte}{\underline{u}ltE}{conj.}{and (connects two clauses)}{S}
\word{ioang}{ioaN}{n.}{animal}{S}
\word{tsap'alute}{\texttslig apP\underline{a}lutE}{n.}{apology}{S}
\word{pxasul}{\underline{p'a}sul}{adj.}{appealing as food}{S}
\word{lam}{lam}{v.}{appear}{S}
\word{pxun}{p'un}{n.}{arm}{S}
\word{p\"ahem}{\underline{p\ae}hEm}{v.}{arrive}{S}
\word{swizaw}{swizaw}{n.}{arrow}{S}
\word{tseo}{\texttslig Eo}{n.}{art}{S}
\word{na}{na}{conj.}{as}{S}
\word{pxel}{p'El}{conj.}{as}{S}
\derive{tengfya}{tENfja}{conj.}{as}{*teng}{the same way as}{*fya}{path}{S}
\word{pawm}{pawm}{v.}{ask}{S}
\word{srung}{sRuN}{n.}{assistance}{S}
\word{Rongloa}{RoNloa}{prop.n}{assumed clan name}{C}
\word{'eko}{P\underline{E}ko}{}{attack}{S}
\word{unilt\`irantokx}{uniltI\underline{Ran}tok'}{n.}{avatar}{S}
\word{txen}{t'En}{adj.}{awake}{S}
\derive{neto}{nEto}{prep.}{away (direction)}{ne}{to}{*to}{away}{S}
\word{m\`iso}{mIso}{prep.}{away (position)}{S}
\derive{*to}{to}{prep.}{away from}{neto}{away (direction)}{m\`iso}{away (direction)}{T}
\end{hangparas} }\end{multicols}

\noindent\textbf{B}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\derives{*\`im}{Im}{prep.}{back}{ne'\`im}{back}{T}
\word{ne'\`im}{nEP\underline{Im}}{prep.}{back (direction)}{S}
\cw{kawng}{kawN}{adj.}{bad}{*kaw--}{not}{--ng}{inclusive}{S} $_T$
\word{p\"ats\`i}{\underline{p\ae}\texttslig I}{n.}{badge}{S}
\derives{*rum}{Rum}{n.}{ball}{rumut}{ball tree}{T}
\word{rumut}{Rumut}{n.}{ball tree (Pandoran flora)}{SG}
\word{utu mauti}{utu mauti}{n.}{banana fruit}{SG}
\word{ikran}{\underline{ik}Ran}{n.}{banshee}{C} ({\it mountain banshee}) $_{SG}$
\word{awaiei}{awaiEi}{n.}{banshee of paradise (Pandoran flora)}{SG}
\word{lu=}{lu}{v.}{be (predicative copula}{S}
\derive{l$<$iv$>$u}{li\underline{vu}}{v.}{be (subjnuctive)}{lu}{be}{$<$iv$>$}{subjunctive infix}{MS}
\word{tsun}{\texttslig un}{v.}{be able}{S}
\word{tok}{tok}{v.}{be at}{S}
\word{nari si}{naRi si}{v.}{be careful}{S}
\word{tam tam}{tam tam}{intj.}{be content}{M}
\word{tsranten}{\texttslig RantEn}{v.}{be of importance}{S}
\word{fnu}{fnu}{v.}{be quiet}{S}
\word{kllfrivo'}{k\textsyllabic{l}:fRi\underline{vo}P}{v.}{be responsible}{S}
\word{ioang}{ioaN}{n.}{beast}{S}
\word{'ekong}{P\underline{E}koN}{}{beat (rhythmic)}{S}
\word{lor}{loR}{adj.}{beautiful (non-human)}{F}
\derive{loreyu}{loREju}{n.}{beautiful spiral (Pandoran Flora)}{lor}{beautiful}{\`ireyu}{spiral}{SG} $_T$
\word{slu}{slu}{v.}{become}{S}
\word{teylu}{tEjlu}{n.}{beetle larva (Pandoran fauna)}{C,S}
\word{eo}{\underline{E}o}{prep.}{before}{S}
\word{sng\"a'i}{sN\ae Pi}{v.}{begin}{S}
\cw{sng\"a'ikrr}{\underline{sN\ae}Pik\textsyllabic{r}}{n.}{beginning}{sng\"a'i}{begin}{krr}{time}{S}
\derives{*\`im}{Im}{prep.}{behind}{ne'\`im}{back}{T}
\word{tsawl}{\texttslig awl}{adj.}{big (in stature)}{S}
\cw{penghrrap}{pENh\textsyllabic{r}ap}{n.}{binary sunshine}{peng}{tell}{hrrap}{danger}{SG}
\word{y\`im}{jIm}{v.}{bind}{S}
\word{'it}{P\underline{i}t}{n.}{bit}{S}
\word{n\`i'it}{nIPit}{n.}{bit}{S}
\word{'ong}{PoN}{v.}{blossom}{S}
\word{ean}{\underline{E}an}{}{blue}{S}
\word{seze}{\underline{sE}zE}{n.}{blue flower (Pandoran flora)}{S}
\word{Omatikaya}{omati\underline{ka}ja}{prop.n.}{Blue Flute Clan}{C,S}
\word{tokx}{\underline{tok'}}{n.}{body}{S, MS}
\word{tsahaylu}{\texttslig a\underline{hEj}lu}{n.}{bond}{C,S} (changed prnounciation for phonetic reasons) $_T$
\word{kxangangang}{\underline{k'aN}aNaN}{n.}{boom (onomatopoeia)}{S}
\word{tsko}{\texttslig ko}{n.}{bow (weapon)}{S}
\word{tsko swizaw}{\texttslig ko swizaw}{n.}{bow and arrows}{S}
\word{eltu}{\underline{El}tu}{n.}{brain}{S}
\cw{eltungawng}{\underline{El}tuNawN}{n.}{brainworm (Pandoran fauna)}{eltu}{brain}{ngawng}{worm}{S} $_T$
\word{vul}{vul}{n.}{branch (of a tree)}{S}
\word{tsteu}{\texttslig etEu}{adj.}{brave}{S} (might be {\bf tstew})$_D$
\word{munge}{\underline{mu}NE}{v.}{bring}{S}
\word{zamunge}{za\underline{mu}NE}{v.}{bring}{S}
\word{tsmukan}{\underline{\texttslig muk}an}{n.}{brother}{S,MS}
\word{nekx}{nEk'}{v.}{burn}{S}
\word{sl\"a}{sl\ae}{conj.}{but}{S}
\word{tx\`im}{t'Im}{n.}{butt}{S}
\word{\`il\"a}{Il\ae] or [\underline{I}l\ae}{prep.}{by}{S}
\word{fa}{fa}{prep.}{by means of}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{C}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{syaw}{sjaw}{v.}{call}{S}
\word{mawey}{mawEj}{adj.}{calm}{S}
\word{tam tam}{tam tam}{intj.}{calm}{M}
\word{tsun}{\texttslig un}{v.}{can}{S}
\cw{spe'etu}{spEP\underline{E}tu}{n.}{captive}{spe'e}{capture}{tu}{person}{S}
\word{spe'e}{spEP\underline{E}}{v.}{capture}{S}
\word{pamtseowll}{pam\texttslig Eow\textsyllabic{l}:}{n.}{cat ear}{SG}
\cw{kxamtseng}{\underline{k'am}\texttslig EN}{n.}{center-place}{kxam}{center}{tseng}{place}{S}
\word{law}{law}{adj.}{certain}{S}
\cw{yomioang}{jomioaN}{n.}{chalice plant (Pandoran flora)}{yom}{eat}{iaong}{animal $_T$}{SG}
\word{fpeio}{fpE\underline{i}o}{}{challenge (ceremonial)}{S}
\word{latem}{\underline{la}tEm}{v.}{change}{S}
\word{kxll}{k'\textsyllabic{l}:}{}{charge}{G}
\word{p\"angkxo}{pANk'o}{v.}{chat}{F}
\word{'eveng}{P\underline{E}vEN}{n.}{child}{S}
\word{Ralu}{\underline{Ra}lu}{prop.n.}{child's name}{S}
\word{ftxey}{ftEj}{v.}{choose}{S}
\word{olo'}{oloP}{n.}{clan}{S}
\cw{olo'eyktan}{\underline{o}loP\underline{Ejk}tan}{n.}{clan leader}{olo'}{clan}{eyktan}{leader}{C,S} (male clan leader) $_{SG}$
\word{Tipnai}{tipnai}{prop.n.}{clan name}{G}
\word{law}{law}{adj.}{clear}{S}
\word{s\`ilronsem}{sIl\underline{Ron}sEm}{adj.}{clever (thing)}{S}
\word{'awkx}{Pawk'}{n.}{cliff}{S}
\word{lok}{lok}{prep.}{close}{S}
\word{lok}{lok}{prep.}{close to}{S}
\word{txewk}{t'Ewk}{n.}{club (weapon)}{G}
\word{lertu}{lERtu}{n.}{colleague}{S}
\word{za'u}{\underline{za}Pu}{v.}{come}{S}
\derive{z$<$er$>$a'u}{ze\underline{ra}Pu}{v.}{come}{za'u}{come}{$<$er$>$}{imperfective aspect}{MS}
\word{to}{to}{conj.}{comparative marker}{S}
\derive{*nul}{nul}{adp.}{comparitive adjectival marker}{n\`i'ul}{more}{nulkrr}{longer}{T,D}
\word{n\`iwotx}{nIwot'}{adv.}{completely}{S}
\word{nekx}{nEk'}{v.}{consume}{S}
\word{n\`itut}{nItut}{adv.}{continually}{S}
\derive{*ran}{Ran}{v.}{contracted form of walk}{t\`iran}{walk}{ikran}{walks-the-heavens}{D}
\word{telem}{tElEm}{n.}{cord}{S} (could be {\it chord}) $_T$
\word{ngop}{Nop}{v.}{create}{S}
\word{swir\"a}{swiR\ae}{n.}{creature}{S}
\word{tskalep}{\texttslig kalEp}{n.}{crossbow}{G}
\derives{*tsngal}{\texttslig Nal}{n.}{cup}{tawtsngal}{sky cup}{T}
\word{meoauniaea}{mEoauniaEa}{}{currently undefined}{F}
\word{mun'i}{munP\underline{i}}{}{cut}{S}
\cw{tsyorina'wll}{\texttslig jorinaPw\textsyllabic{l}:}{n.}{cycad}{rina'}{seed}{*wll}{plant}{SG} $_T$
\end{hangparas} }\end{multicols}

\noindent\textbf{D}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\note{sempu}{sEmpu}{n.}{dad}{affectionate form of}{sempul}{father}{S}
\derive{yomh\`i'ang}{jomhIPaN}{n.}{dakteron (Pandoran flora)}{h\`i'i}{small}{ioang}{animal}{SG} $_T$
\word{sreu}{sREu}{v.}{dance}{S} (might be {\bf srew}) $_D$
\word{hrrap}{h\textsyllabic{r}ap}{n.}{danger}{S}
\word{paywll}{pajw\textsyllabic{l}:}{n.}{dapophet (Pandoran flora)}{SG}
\word{'ite}{P\underline{i}te}{n.}{daughter}{S}
\derive{Mo'at'ite}{moPatPitE}{prop.n.}{daughter of Mo'at}{Mo'at}{woman}{ite}{daughter}{C}
\word{trr}{t\textsyllabic{r}}{n.}{day}{S}
\cw{kerusey}{\underline{kE}RusEj}{adj.}{dead}{ke}{not}{r$<$us$>$ey}{living}{S} $_D$
\word{vrrtep}{\underline{v\textsyllabic{r}}tEp}{n.}{demon}{S}
\word{hum}{hum}{v.}{depart}{S}
\word{ska'a}{skaP\underline{a}}{v.}{destroy}{S}
\word{terkup}{\underline{tER}kup}{v.}{die}{S}
\word{kllkulat}{k\textsyllabic{l}:\underline{ku}lat}{v.}{dig up}{S}
\word{snum\`ina}{\underline{snu}mIna}{adj.}{dim (of a person)}{S}
\word{wutso}{wu\texttslig o}{n.}{dinner}{S}
\derive{*fya--}{fja}{n.}{direction}{fya'o}{path}{fyape}{how}{T}
\word{pa'li}{paPli}{n.}{direhorse (Pandoran fauna)}{S}
\word{tawng}{tawN}{v.}{dive}{S} %eliminated duck n.
\word{si}{si}{v.}{do}{S}
\word{tam}{tam}{v.}{do}{S}
\word{r\"a'\"a}{R\ae P\ae}{v.}{do not}{S}
\loan{toktor}{\underline{tok}toR}{n.}{doctor}{doctor}{S}
\word{hasey}{hasEj}{adj.}{done}{S}
\word{unil}{\underline{u}nil}{n.}{dream}{S}
\word{Uniltaron}{\underline{u}nil\underline{ta}Ron}{prop.n.}{Dream Hunt}{C,S,MS}
\word{unilt\`iranyu}{unitI\underline{Ran}ju}{n.}{dreamwalker}{S}
\word{unilt\`irantokx}{uniltI\underline{Ran}tok'}{n.}{dreamwalker body}{S}
\word{kurakx}{kuRak'}{v.}{drive out}{S}
\word{au}{au}{n.}{drum}{S}
\word{tawng}{tawN}{v.}{duck}{S} %eliminated duck n.
\word{kelku si}{kElku si}{v.}{dwell}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{E}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{mikyun}{\underline{mik}jun}{n.}{ear}{C,S}
\word{n\`iftue}{nI\underline{ftu}E}{adv.}{easily}{S}
\word{ftue}{ftuE}{adj.}{easy}{S}
\word{yom}{jom}{v.}{eat}{S}
\word{pxuntil}{p'until}{n.}{elbow}{S}
\loan{'\`ingl\`is\`i}{P\underline{IN}lIsI}{n.}{English language}{English}{S}
\word{n\`itam}{nItam}{adv.}{enough}{S} %tam
\word{fpx\"ak\`im}{fp'\ae kIm}{v.}{enter}{S}
\cww{pxorna'}{p'ornaP}{n.}{episoth}{rina'}{seed}{SG} $_T$
\word{*krr}{k\textsyllabic{r}}{adv.}{ever}{M} (derived from direct transcription)
\derive{*fra--}{fRa}{adp.}{every}{frapo}{everyone}{fra'u}{everything $_D$}{T}
\cww{frapo}{\underline{fRa}po}{n.}{everyone}{po}{he/she}{S,MS}
\cww{fra'u}{fRaPu}{n.}{everything}{'u}{thing}{S}
\cw{kawng}{kawN}{adj.}{evil}{*kaw--}{not}{--ng}{inclusive}{S} $_T$
\word{t\`ikawng}{tI\underline{kawN}}{n.}{evil}{S,MS}
\word{steftxaw}{stEft'aw}{v.}{examine}{S}
\word{t\`ikenong}{tI\underline{kE}nong}{n.}{example}{S}
\word{n\`ihawng}{nIhawN}{adv.}{excessively}{S}
\word{nang}{naN}{part.}{exclamation}{S}
\word{kenong}{\underline{kE}noN}{v.}{exemplify}{S}
\word{tskxekeng}{\texttslig k'EkEN}{n.}{exercise}{S}
\word{tok}{tok}{v.}{exist}{S}
\word{lu=}{lu}{v.}{existential verb)}{S}
\cww{pxorna'}{p'ornaP}{n.}{exploding seed (Pandoran flora)}{rina'}{seed}{SG} $_T$
\word{nari}{\underline{na}Ri}{n.}{eye}{C,S}
\cw{menari}{mE\underline{na}Ri}{n.}{eyes}{me+}{dual}{nari}{eye}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{F}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{mui\"a}{mui\ae}{adj.}{fair}{S}
\word{tsleng}{\texttslig lEN}{adj.}{false}{S}
\word{al\`im}{alIm}{}{far}{?}
\word{al\`im}{alIm}{}{far away}{?}
\word{n\`iwin}{nIwin}{adv.}{fast}{S}
\word{sempul}{\underline{sEm}pul}{n.}{father}{S}
\word{txopu}{t'opu}{}{fear}{S}
\cw{yomt\`ing}{\underline{jom}tIN}{v.}{feed}{yom}{eat}{t\`ing}{give}{S}
\word{tute}{tut\underline{E}}{n.}{female}{F}
\word{zekw\"a}{zEkw\ae}{n.}{finger}{S}
\word{hasey}{hasEj}{adj.}{finished}{S}
\word{txep}{t'Ep}{n.}{fire}{S}
\word{'awve}{PawvE}{adj.}{first}{S} 
\word{n\`i'awve}{nIP\underline{aw}vE}{adv.}{first}{S}
\word{'engeng}{P\underline{E}NEN}{n.}{floor $_{T}$}{S}
\derives{*tsyo}{\texttslig jo}{n.}{flour}{tsyorina'wll}{flour seed plant}{T}
\cw{tsyorina'wll}{\texttslig jorinaPw\textsyllabic{l}:}{n.}{flour seed plant (Pandoran flora)}{rina'}{seed}{*wll}{plant}{SG} $_T$
\word{tswon}{\texttslig won}{v.}{fly}{S} (Corrected from {\bf tsw$<$ay$>$on}) $_T$
\word{\`il\"a}{Il\ae] or [\underline{I}l\ae}{prep.}{following}{S}
\word{fpi}{fpi}{}{for the sake of}{S}
\word{kxan\`i}{k'anI}{adj.}{forbidden}{S}
\word{na'r\`ing}{\underline{na}PRIN}{n.}{forest}{S}
\word{ikranay}{ikRanaj}{n.}{forest banshee}{SG} (dubious: +ay would make {\it banshees}) $_T$
\word{txoa}{t'oa}{n.}{forgiveness}{S}
\word{ts\`ing}{\texttslig IN}{num.}{four}{S}
\word{pxasul}{\underline{p'a}sul}{adj.}{fresh}{S}
\cw{penghrrap}{pENh\textsyllabic{r}ap}{n.}{fringed lamp (Pandoran flora)}{peng}{tell}{hrrap}{danger}{SG}
\word{ftu+}{ftu}{prep.}{from (direction)}{S}
\word{ta}{ta}{prep.}{from (various uses)}{S}
\word{ta'em}{taPEm}{prep.}{from above}{S}
\word{meoauniaea}{mEoauniaEa}{}{Frommer's favourite word}{F}
\word{teya}{tEja}{adj.}{full}{S}
\word{hiy\`ik}{\underline{hi}jIk}{adj.}{funny}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{G}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{sngel}{sngEl}{n.}{garbage}{F}
\cw{sngeltseng}{sngEl\texttslig EN}{n.}{garbage place}{sngel}{garbage}{tseng}{place}{F}
\cw{ne kllte}{nE \underline{k\textsyllabic{l}:}tE}{intj.}{get down}{ne}{to}{kllte}{the ground}{S}
\word{t\`ing}{tIN}{v.}{give}{S}
\word{k\"a}{k\ae}{v.}{go}{S}
\word{saleu}{sa\underline{le}u}{v.}{go}{MS} (might be from {\bf za'u}) $_T$
\word{s\`iltsan}{sIl\texttslig an}{adj.}{good}{S}
\word{Ninat}{nInat}{prop.n.}{good singer}{C}
\cww{k\`iyevame}{kIjE\underline{va}mE}{v.}{good-bye}{kame}{see}{S}
\derives{tslolam}{\texttslig lolam}{intj.}{got it}{tsl$<$ol$>$am}{understand}{S} with perfective aspect infix $_T$
\word{ni\"a}{ni\ae}{v.}{grab}{S}
\derives{teswot\`ing}{tE\underline{swo}tIN}{v.}{grant}{t\`ing}{give}{S}
\word{nawm}{nawm}{adj.}{great}{S}
\word{txan}{t'an}{adj.}{great}{S}
\word{txan-}{\underline{txan}}{adj.}{great}{MS}
\word{toruk}{\underline{to}Ruk}{n.}{great leonopteryx}{C,S}
\word{torukspxam}{toRuksp'am}{n.}{great leonopteryx fungus (Pandoran flora)}{SG}
\derive{*kll--}{k\textsyllabic{l}:}{}{ground}{kllkulat}{dig up}{kllkxem}{stand}{T}
\word{kllte}{k\textsyllabic{l}:tE}{n.}{ground}{S}
\word{pongu}{\underline{po}Nu}{pn.}{group}{S}
\cw{fyaw\`intxu}{fjawIn\underline{t'u}}{}{guide}{*fya}{how}{w\`intxu}{show}{S} $_T$
\word{Eywa}{\underline{Ej}wa}{prop.n.}{guiding force and deity (equivalent to Gaia)}{C} \textbf{\underline{Ey}wa \underline{nga}hu} intj. {\it good-bye, ... be with you} $_S$
\loan{kuns\`ip}{\underline{kun}sIp}{n.}{gunship}{gunship}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{H}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{txey}{t'Ej}{v.}{halt}{S}
\word{'angts\`ik\"a}{PaN\texttslig Ik\ae}{n.}{hammerhead (Pandora fauna)}{G}
\derives{*lrr}{l\textsyllabic{r}}{}{happy}{lrrtok}{smile exists}{T}
\word{po}{po}{pn.}{he}{S}
\word{poan}{poan}{pn.}{he}{S}
\word{txe'lan}{t'EP\underline{lan}}{n.}{heart}{S,MS}
\derive{*ik}{ik}{n.}{heaven}{ik-ni-maya}{stairway to heaven}{ikran}{walks the sky}{D}
\derive{loreyu}{loREju}{n.}{helicoradium spirale}{lor}{beautiful}{\`ireyu}{spiral}{SG} $_T$
\word{kaltx\`i}{kalt'I}{intj.}{hello}{S}
\word{srung}{sRuN}{n.}{help}{S}
\word{f\`itseng(e)}{fI\underline{\texttslig EN}(E)}{pn.}{here}{S,MS}
\word{yerik}{\underline{jE}Rik}{n.}{hexapede (Pandoran fauna)}{S}
\word{Tsah\`ik}{\underline{\texttslig a}hIk}{prop.n.}{high priestess}{S}
\word{til}{til}{n}{hinge}{S}
\word{fpak}{fpak}{v.}{hold off}{S}
\word{ftxoz\"a}{ft'oz\ae}{n.}{holiday}{FE}
\word{kelku}{\underline{kEl}ku}{n.}{home}{S}
\word{Kelutral}{\underline{kE}lutRal}{prop.n.}{Hometree (Pandoran flora)}{S,MS}
\word{meuia}{mE\underline{u}ia}{}{honour}{S}
\word{txll'u}{t'\textsyllabic{l}:Pu}{n.}{hookagourd (Pandoran flora)}{SG}
\word{som}{som}{adj.}{hot}{S}
\cw{somt\`ilor}{somtIlor}{n.}{hot beauty (Pandoran flora)}{som}{hot}{t\`ilor}{beauty}{SG}
\word{fyape}{fjapE}{conj.}{how}{S}
\word{pefya}{pEfja}{adv.}{how}{S}
\word{nguway}{Nuwaj}{n.}{howl}{G}
\word{taron}{\underline{ta}Ron}{v.}{hunt}{S}
\cw{taronyu}{\underline{ta}Ronju}{n.}{hunter}{taron}{hunt}{--yu}{agent suffix}{C,S}
\end{hangparas} }\end{multicols}

\noindent\textbf{I}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{oe}{\underline{o}E}{pn.}{I, me}{S}
\word{oe-}{\underline{oe}-}{pn.}{I, me}{MS}
\word{ohe-}{o\underline{he}-}{pn.}{I, me (deferential or ceremonial form)}{MS}
\word{ohe}{\underline{o}hE}{pn.}{I, me (deferential or ceremonial form)}{S, MS}
\derives{tslolam}{\texttslig lolam}{intj.}{I understand}{tsl$<$ol$>$am}{understand}{S} with perfective aspect infix $_T$
\word{txo}{t'o}{conj.}{if}{S}
\cww{txokefyaw}{\underline{t'ok}Efjaw}{conj.}{if not}{txo-ke-fya-aw}{if not one path}{S}
\word{m\`i+}{mI}{prep.}{in}{S}
\word{eo}{\underline{E}o}{prep.}{in front of}{S}
\word{ha}{ha}{conj.}{in that case}{S}
\word{n\`iwotx}{nIwot'}{adv.}{in toto}{S}
\word{Na'vi}{\underline{na}Pvi}{prop.n.}{indigenous Pandoran sentient race}{C,S,MS}
\word{keye'ung}{kE\underline{jE}PuN}{n.}{insanity}{S}
\word{s\"anume}{s\ae\underline{nu}mE}{n.}{instruction}{S}
\word{zoplo}{zop\underline{lo}}{n.}{insult}{S}
\word{ralpeng}{RalpEN}{v.}{interpret}{S}
\word{Tsah\`ik}{\underline{\texttslig a}hIk}{prop.n.}{interpreter}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{J}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{til}{til}{n}{joint}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{K}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\note{'evi}{P\underline{E}vi}{n.}{kid}{affectionate form of}{'eveng}{child}{S}
\word{tspang}{\texttslig paN}{v.}{kill}{S}
\word{k\`i'ong}{kIPoN}{n.}{kind of fruit or vegetable (Pandoran flora)}{S}
\word{kxener}{\underline{k'E}nER}{n.}{kind of fruit or vegetable (Pandoran flora)}{S}
\word{kinamtil}{ki\underline{nam}til}{n.}{knee}{S}
\word{kame}{\underline{ka}mE}{v.}{know}{S}
\word{omum}{\underline{o}mum}{v.}{know}{S}
\end{hangparas} }\end{multicols}
\newpage
\noindent\textbf{L}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{atxkxe}{at'k'E}{n.}{land}{S}
\derive{l\`i'fya}{lIfja}{n.}{language}{l\`i'u}{word}{fya}{way $_W$}{F}
\word{apxa}{ap'a}{adj.}{large}{S}
\word{toruk}{\underline{to}Ruk}{n.}{last shadow}{C,S}
\word{eyk}{Ejk}{v.}{lead}{S}
\word{eyktan}{\underline{Ejk}tan}{n.}{leader}{S}
\word{nume}{numE}{v.}{learn}{S}
\word{hum}{hum}{v.}{leave}{S}
\word{tx\`ing}{t'IN}{v.}{leave}{S}
\word{kinam}{kinam}{n.}{leg}{S}
\cw{fteke}{ftEkE}{conj.}{lest}{fte}{so that}{ke}{not}{S}
\word{lonu}{lonu}{v.}{let go}{S}
\word{'engeng}{P\underline{E}NEN}{n.}{level}{S}
\word{t\`irey}{tIREj}{n.}{life}{S}
\word{atan}{atan}{}{light}{S}
\word{na}{na}{conj.}{like}{S}
\word{pxel}{p'El}{conj.}{like}{S}
\derive{*teng--}{tEN}{adp.}{like}{tengfya}{same way as}{tengkrr}{while}{T}
\word{f\`ifya}{fIfja}{pn.}{like this}{S} %fja'o
\word{n\`iayoeng}{nIaojoEN] or [naj\underline{wEN}}{adv.}{like us}{S} (corrected from {\bf -eg}) $_T$
\word{seyri}{sEjRi}{n.}{lip}{C,S}
\word{eltu si}{\underline{El}tu si}{v.}{listen}{S}
\word{t\`ing mikyun}{tIm \underline{mik}jun}{v.}{listen}{S}
\word{t\`ing mikyun}{tIm \underline{mik}jun}{v.}{lit. give an ear}{S}
\word{t\`ing nari}{tIn \underline{na}Ri}{v.}{lit. give an eye}{S}
\word{tsawlapxangrr}{\texttslig awlap'aN\textsyllabic{r}}{n.}{lit. tall large root (Pandoran flora)}{SG} (for short, {\bf tsawlapx})
\word{kelku si}{kElku si}{v.}{live}{S}
\word{rey}{REj}{v.}{live}{S}
\word{sutx}{sut'}{v.}{lock up}{S} % should be trap?
\word{ngim}{Nim}{adj.}{long}{F}
\word{yol}{jol}{adj.}{long (period of time)}{S}
\word{nulkrr}{\underline{nu}lk\textsyllabic{r}}{}{longer (time)}{S} %nul?
\word{n\`in}{nIn}{v.}{look}{S}
\word{t\`ing nari}{tIn \underline{na}Ri}{v.}{look}{S}
\word{n\`in}{nIn}{v.}{look at}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{M}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{si}{si}{v.}{make}{S}
\word{tutean}{tutE\underline{an}}{n.}{male}{S}
\word{Eytukan}{\underline{Ej}tukan}{prop.n.}{man's name}{C,MS}
\word{Tsu'tey}{\texttslig uP\underline{tEj}}{prop.n.}{man's name}{C}
\word{pxay}{p'aj}{adj., pn.}{many}{S}
\word{txan}{t'an}{adj.}{many}{S}
\word{muntxa}{munt'a}{adj.}{mated}{S}
\word{Tsah\`ik}{\underline{\texttslig a}hIk}{prop.n.}{matriarch}{S}
\word{tsranten}{\texttslig RantEn}{v.}{matter}{S}
\word{txele}{t'ElE}{v.}{matter (subject)}{S}
\word{kxawm}{k'awm}{adv.}{maybe}{S}
\word{oe}{\underline{o}E}{pn.}{me}{S}
\word{ral}{Ral}{n.}{meaning}{S}
\word{fpxafaw}{fp'afaw}{n.}{medusa (Pandoran fauna)}{S}
\word{lonataya}{lonataja}{n.}{medusa (Pandoran Fauna)}{SG}
\word{ultxa}{ult'a}{n.}{meeting}{S}
\word{'upxare}{Pup'aRE}{}{message}{S}
\word{fngap}{fngap}{n.}{metal}{F}
\word{fngapsutxwll}{fngapsut'w\textsyllabic{l}:}{n.}{metal-following plant (Pandoran flora)}{SG}
\word{kxam}{k'am}{n.}{middle}{S}
\word{kxam}{k'am}{n.}{midpoint}{S}
\word{fkeu}{fkEu}{adj.}{mighty}{S} (might be {\bf fkew}) $_D$
\word{ronsem}{\underline{Ron}sEm}{n.}{mind}{S}
\word{kenong}{\underline{kE}noN}{v.}{model}{S}
\word{swaw}{swaw}{n.}{moment}{S}
\word{n\`i'ul}{nIPul}{adv.}{more}{S} 
\word{skxawng}{sk'awng}{n.}{moron}{S}
\word{sa'nok}{saPnok}{n.}{mother}{S} (from {\bf sa'atenuk} $_C$)
\derive{*ik}{ik}{n.}{mountain}{ik-ni-maya}{stairway to heaven}{ikran}{walks the sky}{D}
\word{rikx}{Rik'}{v.}{move}{S}
\derives{*win}{win}{n.}{move swiftly}{n\`iwin}{fast}{T}
\word{txan}{t'an}{adj.}{much}{S}
\word{n\`itxan}{nIt'an}{adv.}{much (abstract)}{S}
\note{sa'nu}{saPnu}{n.}{mum}{affectionate form of}{sa'nok}{mother}{S}
\derives{*spxam}{sp'am}{n.}{mushroom}{torukspxam}{octoshroom}{T}
\derive{pamtseo}{\underline{pam}\texttslig Eo}{n.}{music}{pam}{music}{tseo}{art}{S}
\word{pamtseowll}{pam\texttslig Eow\textsyllabic{l}:}{n.}{music plant (Pandoran flora)}{SG}
\word{zene}{zEnE}{v.}{must}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{N}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{kin}{kin}{}{need}{S}
\word{t\`ikin}{tIkin}{n.}{need}{S}
\cww{kawkrr}{\underline{kaw}k\textsyllabic{r}}{adv.}{never}{krr}{time}{S}
\word{fmawn}{fmawn}{n.}{news}{S}
\word{txon}{t'on}{n.}{night}{S}
\word{ke--}{kE}{adp.}{no}{T}
\word{kehe}{\underline{kE}hE}{}{no}{S}
\cww{kawtu}{kawtu}{pn.}{no-one}{tu}{person}{S}
\word{nawm}{nawm}{adj.}{noble}{S}
\word{faketuan}{fakEtuan}{n.}{non-Na'vi}{?} (might be poor transcription of {\bf fa} + {\bf ketuan}) $_{T}$
\word{ontu}{\underline{o}ntu}{n.}{nose}{C,S}
\word{ke}{kE}{adv., adj.}{not}{S}
\word{ke--}{kE}{adp.}{not}{T}
\cw{ke'u}{\underline{kE}Pu}{pn.}{nothing}{ke}{not}{'u}{thing}{S}
\word{set}{sEt}{adv.}{now}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{O}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\derives{*p\"anu}{p\ae nu}{n.}{oath}{p\"anut\`ing}{promise}{D}
\word{tok}{tok}{v.}{occupy}{S}
\word{torukspxam}{toRuksp'am}{n.}{octoshroom}{SG}
\word{zoplo}{zop\underline{lo}}{n.}{offence}{S}
\word{aungia}{auNia}{n.}{omen}{S}
\word{'aw}{Paw}{num.}{one}{S}
\word{'aw--}{Paw}{pn.}{one}{T}
\word{fko}{fko}{}{one}{S} (might be from {\bf po}) $_{T}$ 
\word{'awpo}{Pawpo}{pn.}{one individual}{S} $_{T}$
\word{n\`i'aw}{nIPaw}{adj.}{only}{S}
\word{fu}{fu}{conj.}{or}{S}
\cww{txokefyaw}{\underline{t'ok}Efjaw}{conj.}{or else}{txo-ke-fya-aw}{if not one path}{S}
\derive{*la}{la}{adp.}{other}{lapo}{other one}{aylahe}{others}{T}
\word{lahe}{\underline{la}hE}{pn.}{other}{S}
\cww{lapo}{\underline{la}po}{pn.}{other one}{po}{he/she}{S}
\derive{*la}{la}{adp.}{otherness}{lapo}{other one}{aylahe}{others}{T}
\word{wrrpa}{w\textsyllabic{r}pa}{adj.}{outside}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{P}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\cww{tawtsngal}{taw\texttslig Nal}{n.}{panopyra (Pandoran flora)}{taw}{sky}{SG}
\word{hapx\`i}{ha\underline{p'I}}{}{part}{S,MS}
\word{pongu}{\underline{po}Nu}{pn.}{party}{S}
\derive{emza'u}{Emz\underline{a}Pu}{v.}{pass (a test)}{em}{over}{za'u}{come}{S,MS}
\derives{*ftxavang}{ft'avaN}{adj.}{passion}{n\`iftxavang}{passionately}{T}
\word{n\`iftxavang}{nI\underline{ft'a}vaN}{adv.}{passionately}{S}
\derive{*fya--}{fja}{n.}{path}{fya'o}{path}{fyape}{how}{T}
\word{fya'o}{fjaPo}{n.}{path}{S}
\word{renu}{\underline{RE}nu}{n.}{pattern}{S}
\word{eltu si}{\underline{El}tu si}{v.}{pay attention}{S}
\word{fpom}{fpom}{}{peace}{S}
\word{fkarut}{\underline{fka}Rut}{}{peel}{S}
\word{kxawm}{k'awm}{adv.}{perhaps}{S}
\word{*tu}{tu}{n.}{person}{T}
\word{tute}{\underline{tu}tE}{n.}{person}{F}
\word{*tu}{tu}{n.}{personage marker}{T}
\word{tseng(e)}{\texttslig EN(E)}{n.}{place}{S}
\word{tei}{tEi}{n.}{plains}{G}
\derive{*wll}{w\textsyllabic{l}:}{n.}{plant}{pamtseowll}{music plant}{txumts\"a'wll}{poison-squirting plant}{T}
\word{rutxe}{Rut'E}{v.}{please}{S}
\word{prrte'}{p\textsyllabic{r}tEP}{n.}{pleasure}{F}
\word{txum}{t'um}{n.}{poison}{S}
\cw{txumpaywll}{t'umpayw\textsyllabic{l}:}{n.}{poison water plant (Pandoran flora)}{txum}{poison}{paywll}{water plant}{SG}
\cw{somt\`ilor}{somtIlor}{n.}{popsicle}{som}{hot}{t\`ilor}{beauty}{SG}
\word{sevin}{sEvin}{adj.}{pretty}{S} (only for people) $_F$
\derive{*'it}{Pit}{n.}{progeny}{'itan}{son}{'ite}{daughter}{T}
\word{vir\"a}{viR\ae}{v.}{proliferate}{S} (possible loan from English {\it virus}) $_T$
\derives{*p\"anu}{p\ae nu}{n.}{promise}{p\"anut\`ing}{promise}{D}
\derives{p\"anut\`ing}{\underline{p\ae}nutIN}{v.}{promise}{t\`ing}{give}{S}
\word{mui\"a}{mui\ae}{adj.}{proper}{S}
\word{hawnu}{hawnu}{v.}{protect}{S}
\word{t\`ihawnu}{tI\underline{haw}nu}{n.}{protection}{S}
\word{za'\"ar\`ip}{zaP\underline{\ae}RIp}{v.}{pull}{S}\
\word{utu mauti}{utu mauti}{n.}{push fruit (Pandoran flora)}{SG}
\word{utral utu mauti}{utral utu mauti}{n.}{push fruit tree (Pandoran flora)}{SG}
\end{hangparas} }\end{multicols}

\noindent\textbf{Q}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{fnu}{fnu}{v.}{quiet}{S}
\word{san}{san}{n.}{quote}{S} (works with s\`ik) $_T$
\end{hangparas} }\end{multicols}

\noindent\textbf{R}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{tompa}{tompa}{n.}{rain}{S}
\word{pxiut}{p'iut}{n.}{razor palm}{SG}
\word{alaks\`i}{al\underline{ak}sI}{adj.}{ready}{S}
\word{tx\`im}{t'Im}{n.}{rear end}{S}
\derive{*lun}{lun}{n.}{reason}{lumpe}{why}{pelun}{why}{T}
\word{txumts\"a'wll}{txum\texttslig\ae Pw\textsyllabic{l}:}{n.}{reclining flask (Pandoran flora)}{SG}
\cw{zongtseng}{zoN\texttslig EN}{n.}{refuge}{zong}{save}{tseng}{place}{S}
\word{lonu}{lonu}{v.}{release}{S}
\word{'\`i'awn}{PI\underline{Pawn}}{v.}{remain}{S, MS}
\word{'ok}{Pok}{n.}{remembrance}{S}
\word{kenong}{\underline{kE}noN}{v.}{represent}{S}
\word{\"atx\"ale}{\ae \underline{t'\ae}lE}{}{request}{S, MS}
\word{kato}{\underline{ka}to}{n.}{rhythm}{S}
\word{makto}{makto}{v.}{ride}{S}
\cw{k\"amakto}{k\ae\underline{mak}to}{v.}{ride out}{k\"a}{go}{makto}{ride $_T$}{S}
\word{mui\"a}{mui\ae}{adj.}{right}{S}
\word{malmoks}{malmoks}{n.}{rings of stone}{G}
\word{kilvan}{kilvan}{n.}{river}{F}
\word{tskxe}{\texttslig k'E}{n.}{rock}{S}
\word{tul}{tul}{v.}{run}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{S}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{swok}{swok}{adj.}{sacred}{S}
\word{swotu}{swotu}{n.}{sacred place}{S}
\word{swotulu}{swotulu}{n.}{sacred river}{G}
\derives{*atoki}{atoki}{n.}{sacred tree (Pandoran flora)}{atokirina'}{seeds of the great tree}{T}
\cw{zongtseng}{zoN\texttslig EN}{n.}{safe place}{zong}{save}{tseng}{place}{S}
\derive{*teng--}{tEN}{adp.}{same as}{tengfya}{same way as}{tengkrr}{while}{T}
\cww{teng'u}{tENPu}{pn.}{same thing}{'u}{thing}{?}
\cww{tengkrr}{tEN\underline{k\textsyllabic{r}}}{conj.}{same time as}{krr}{time}{S}
\derive{tengfya}{tENfja}{conj.}{same way as}{*teng}{the same way as}{*fya}{path}{S}
\derive{*wong}{woN}{n.}{sameness}{kewong}{alien}{ketuwong}{alien}{T}
\word{zong}{zoN}{v.}{save}{S}
\word{san}{san}{n.}{saying}{S} (works with s\`ik) $_T$
\cw{txumpaywll}{t'umpayw\textsyllabic{l}:}{n.}{scorpion thistle}{txum}{poison}{paywll}{water plant}{SG}
\word{pxasik}{p'asik}{intj.}{screw that!}{F}
\word{kame}{\underline{ka}mE}{v.}{see}{S}
\word{tse'a}{\texttslig EPa}{v.}{see (physical sense)}{S}
\word{kame}{\underline{ka}mE}{v.}{see into}{S}
\cww{k\`iyevame}{kIjE\underline{va}mE}{v.}{see you soon}{kame}{see}{S}
\word{rina'}{RinaP}{n.}{seed}{S}
\cww{atokirina'}{atokiRi\underline{na}P}{n.}{seeds of the great tree}{rina'}{seed}{C,S}
\word{lam}{lam}{v.}{seem}{S}
\word{fpe'}{fpEP}{v.}{send}{S}
\word{wutso}{wu\texttslig o}{n.}{served meal}{S}
\word{k\`ite'e}{kI\underline{tE}PE}{}{service}{S}
\word{kin\"a}{\underline{ki}n\ae}{num.}{seven}{S}
\word{pxi}{p'i}{adj.}{sharp}{S}
\word{pxiut}{p'iut}{n.}{sharp tree (Pandoran flora)}{SG}
\word{po}{po}{pn.}{she}{S}
\word{poe}{po\underline{E}}{pn.}{she}{S}
\word{hawnu}{hawnu}{v.}{shelter}{S}
\word{rikx}{Rik'}{v.}{shift position}{S}
\derive{*s\`ip}{sIp}{n.}{ship}{taws\`ip}{sky-ship}{kuns\`ip}{gunship}{T}
\word{w\`intxu}{wInt'u}{v.}{show}{S}
\word{zawng}{zawN}{n.}{shriek}{G}
\word{tsmuk}{\texttslig muk}{n.}{sibling}{S}
\word{tsmuktu}{\underline{\texttslig muk}tu}{n.}{sibling}{S} (allomorph of {\bf tsmuk} {\it sibling}) $_T$
\word{aungia}{auNia}{n.}{sign}{S}
\word{rol}{Rol}{v.}{sing}{S}
\word{tsmuke}{\texttslig mukE}{n.}{sister}{S}
\word{vofu}{vofu}{num.}{sixteen}{S}
\derive{*ik}{ik}{n.}{sky}{ik-ni-maya}{stairway to heaven}{ikran}{walks the sky}{D}
\word{taw}{taw}{n.}{sky}{F}
\cww{tawtsngal}{taw\texttslig Nal}{n.}{sky cup (Pandoran flora)}{taw}{sky}{SG}
\derive{tautral}{tautRal}{n.}{sky tree (Pandoran flora)}{taw}{sky}{utral}{tree}{SG} $_T$
\cw{Tawtute}{\underline{taw}tutE}{prop.n.}{Sky-person}{taw}{sky}{tute}{person}{S,MS}
\loan{taws\`ip}{tawsIp}{n.}{sky-ship}{ship}{F}
\word{hahaw}{hahaw}{v.}{sleep}{F}
\word{h\`i'i}{\underline{hI}Pi}{adj.}{small}{S}
\word{n\`i'it}{nIPit}{n.}{small amount (abstract)}{S}
\derive{yomh\`i'ang}{jomhIPaN}{n.}{small animal/insect eater}{h\`i'i}{small}{ioang}{animal}{SG} $_T$
\word{fahew}{fa\underline{hEw}}{v.}{smell}{F}
\derives{*lrr}{l\textsyllabic{r}}{}{smile}{lrrtok}{smile exists}{T}
\word{lrrtok}{\underline{l\textsyllabic{r}}tok}{}{smile}{S}
\word{ha}{ha}{conj.}{so}{S}
\word{fte}{ftE}{conj.}{so that}{S}
\cw{fteke}{ftEkE}{conj.}{so that not}{fte}{so that}{ke}{not}{S}
\word{ko}{ko}{part.}{solicit agreement}{FE}
\word{fmawn}{fmawn}{n.}{something to report}{S}
\word{'itan}{Pitan}{n.}{son}{S}
\word{t\`irol}{tIRol}{n.}{song}{S}
\word{way}{waj}{n.}{song}{S}
\word{waytelem}{\underline{waj}tElEm}{n.}{song chord}{S}
\word{ye'r\`in}{\underline{jE}PRIn}{adv.}{soon}{S,MS}
\word{pam}{pam}{n.}{sound}{S}
\derive{*l\`i(')}{lI(P)}{}{speak}{l\`i'u}{word}{l\`i'fya}{language}{W}
\word{plltxe}{p\textsyllabic{l}:t'E}{v.}{speak}{S}
\word{tukru}{tuk\underline{Ru}}{n.}{spear}{S}
\word{kali'weya}{kaliPwEja}{n.}{species of poisonous arachnid (Pandoran fauna)}{SG}
\word{k\`im}{kIM}{v.}{spin}{S}
\word{'\`iheyu}{PIhEju}{}{spiral}{SG}
\word{tirea}{ti\underline{RE}a}{n.}{spirit}{S}
\cw{tireaioang}{tiR\underline{E}ai\underline{o}aN}{n.}{spirit animal}{tirea}{spirit}{ioang}{animal}{S}
\cw{tireafya'o}{tiR\underline{E}a\underline{fja}Po}{n.}{spirit path}{tirea}{spirit}{fya'o}{path}{S}
\word{vir\"a}{viR\ae}{v.}{spread}{S} (possible loan from English {\it virus}) $_T$
\word{Iknimaya}{\underline{ik}ni\underline{ma}ja}{prop.n.}{Stairway to Heaven}{C,S}
\word{kllkxem}{k\textsyllabic{l}:k'Em}{v.}{stand}{S}
\word{tanh\`i}{tan\underline{hI}}{n.}{star}{S}
\word{sng\"a'i}{sN\ae Pi}{v.}{start}{S}
\cw{sng\"a'ikrr}{\underline{sN\ae}Pik\textsyllabic{r}}{n.}{starting time}{sng\"a'i}{begin}{krr}{time}{S}
\word{'\`i'awn}{PI\underline{Pawn}}{v.}{stay}{S, MS}
\word{riti}{Riti}{n.}{stingbat (Pandoran fauna)}{S}
\word{tskxe}{\texttslig k'E}{n.}{stone}{S}
\word{ayskxe}{aj\underline{sk'E}}{n.}{stones}{MS}
\word{ftang}{ftaN}{v.}{stop}{S}
\word{txey}{t'Ej}{v.}{stop}{S}
\word{t\`iftang}{tIftaN}{n.}{stopping}{S}
\word{yey}{jEj}{adj.}{straight}{S}
\word{hiy\`ik}{\underline{hi}jIk}{adj.}{strange}{S}
\word{u'imi}{uPimi}{n.}{streamers}{SG}
\word{t\`itxur}{tIt'uR}{n.}{strength}{S}
\word{takuk}{\underline{ta}kuk}{v.}{strike}{SG}
\word{t\`itakuk}{titakuk}{n.}{strike}{G}
\word{i'en}{iPEn}{n.}{stringed instrument}{S}
\word{txur}{t'uR}{adj.}{strong}{S}
\word{ftia}{fti\underline{a}}{}{study}{F}
\cww{talioang}{\underline{tal}i\underline{o}aN}{n.}{sturmbeest (Pandoran fauna)}{ioang}{animal}{S}
\word{tam}{tam}{v.}{suffice}{S}
\word{tsawke}{\texttslig awkE}{n.}{sun}{S}
\word{nang}{naN}{part.}{surprise}{S}
\word{fpak}{fpak}{v.}{suspend action}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{T}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{kxetse}{\underline{k'E}\texttslig E}{n.}{tail}{S}
\word{munge}{\underline{mu}NE}{v.}{take}{S}
\word{krrnekx}{k\textsyllabic{r}nEk'}{v.}{take/consume time}{S}
\word{tsawl}{\texttslig awl}{adj.}{tall}{S}
\derives{*kar}{kar}{v.}{teach}{karyu}{teacher}{T}
\word{karyu}{kaRju}{n.}{teacher}{S}
\word{s\"anume}{s\ae\underline{nu}mE}{n.}{teaching}{S}
\word{peng}{pEN}{v.}{tell}{S}
\word{kllpx\`iltu}{k\textsyllabic{l}:\underline{p'Il}tu}{n.}{territory}{S}
\word{fmetok}{fm\underline{E}tok}{v.}{test}{S}
\word{t\`ifmetok}{tIf\underline{mE}tok}{n.}{test}{S, MS}
\word{to}{to}{conj.}{than}{S}
\word{palulukan}{\underline{pa}lu\underline{lu}kan}{n.}{Thanator}{S}
\word{irayo}{iRajo}{n., intj.}{thank you}{C,S} (also {\bf ireiyo}) $_{MS}$
\word{a}{a}{pn.}{that}{S}
\derive{*tsa--}{\texttslig a}{adp.}{that}{tsakem}{that action}{tsakrr}{that time}{T}
\word{tsat}{\texttslig at}{pn.}{that (as object)}{S}
\word{tsn\`i}{\texttslig nI}{}{that (function word)}{S}
\derive{futa}{futa}{conj.}{that (subjunctive marker)}{f\`iu-t}{this accusative}{a}{that}{FE}
\cw{tsa'u}{\texttslig aPu}{pn.}{that (thing)}{*tsa}{that}{*'u}{thing}{S} $_T$
\cw{tsatu}{\texttslig atu}{pn.}{that person}{*tsa}{that}{tu}{person}{S}
\cw{tsatseng}{\texttslig a\texttslig EN}{pn.}{that place}{*tsa}{that}{tseng}{place}{S}
\word{Na'vi}{\underline{na}Pvi}{prop.n.}{the People}{C,S,MS}
\word{Iknimaya}{\underline{ik}ni\underline{ma}ja}{prop.n.}{the rite of passage for young Na'vi}{C,S}
\derive{*wong}{woN}{n.}{the same}{kewong}{alien}{ketuwong}{alien}{T}
\cw{tsatseng}{\texttslig a\texttslig EN}{pn.}{there}{*tsa}{that}{tseng}{place}{S}
\word{tam tam}{tam tam}{intj.}{there there}{M}
\cw{fayvrrtep}{fajv\textsyllabic{r}tEp}{n.}{these demons}{fay}{this plural}{vrrtep}{demon}{F} $_T$
\word{ayfo}{ayfo}{pn.}{they}{S}
\derive{fo}{fo}{pn.}{they}{ay+}{deleted plural leniting morpheme}{po}{he/she}{S} $_T$ %make normal text?
\cw{mefo}{\underline{mE}fo}{pn.}{they}{me+}{dual marker}{po}{he/she}{S}
\word{piy\"ahem}{pi\underline{y\ae}hEm}{v.}{they will be?}{T,MS} 
\derive{*'u}{Pu}{n.}{thing}{f\`i'u}{this thing}{tsa'u}{that thing}{T}
\word{fp\`il}{fpIl}{v.}{think}{FE}
\word{ts\`ivol}{\underline{\texttslig I}vol}{num.}{thirty-two thirty-two}{S}
\derive{f\`i=}{fI}{adp.}{this}{f\`ipo}{this one}{f\`i'u}{this thing}{F}
\word{f\`ifya}{fIfja}{pn.}{this}{S} %fja'o
\word{f\`itseng(e)}{fI\underline{\texttslig EN}(E)}{pn.}{this}{S,MS}
\word{f\`i'u}{fIPu}{pn.}{this}{S}
\word{f\`ikem}{fIkEm}{pn.}{this (action)}{S}
\word{f\`ipo}{\underline{fI}po}{pn.}{this one (person or thing)}{S}
\word{f\`itseng(e)}{fI\underline{\texttslig EN}(E)}{pn.}{this place}{S,MS}
\word{f\`i'u}{fIPu}{pn.}{this thing}{S}
\word{f\`ifya}{fIfja}{pn.}{this way}{S} %fja'o
\cw{mefo}{\underline{mE}fo}{pn.}{those two}{me+}{dual marker}{po}{he/she}{S}
\word{k\`ing}{kIN}{}{thread}{S}
\word{krr}{k\textsyllabic{r}}{n.}{time}{S}
\word{*krr}{k\textsyllabic{r}}{adv.}{time}{M} (derived from direct transcription)
\word{ne}{nE}{prep.}{to}{S}
\word{'awsiteng}{Pawsi\underline{tEN}}{adv.}{together}{S}
\word{n\`ihawng}{nIhawN}{adv.}{too}{S}
\derive{toruk makto}{toRuk makto}{n.}{toruk rider}{toruk}{leonopteryx}{makto}{ride}{C}
\word{'ampi}{Pampi}{}{touch}{S}
\word{ne}{nE}{prep.}{towards}{S}
\word{sutx}{sut'}{v.}{track}{S} % should be trap?
\word{tskxekeng}{\texttslig k'EkEN}{n.}{training}{S}
\derive{*--ut}{ut}{n.}{tree}{pxiut}{sharp tree}{rumut}{ball tree}{SG} (compound-word allomorph of {\bf utral}) $_T$
\word{utral}{utRal}{n.}{tree}{S}
\word{Utral Aymokriy\"a}{\underline{u}tRal aj\underline{mok}Rij\ae}{prop.n.}{Tree of Voices (Pandoran flora)}{S} (from {\bf utraya mokri} $_C$)
\word{ngay}{Naj}{adj.}{true}{S}
\word{n\`ingay}{nINaj}{adv.}{truly}{S} %pronounce?
\word{tangek}{taNEk}{n.}{trunk (of a tree)}{S}
\word{t\`ingay}{tINaj}{n.}{truth}{S}
\word{fmi}{fmi}{v.}{try}{S}
\word{m\`in}{mIn}{}{turn}{S}
\word{mun(e)}{\underline{mu}nE}{num.}{two}{S} (possible as {\bf mun}) $_T$
\end{hangparas} }\end{multicols}

\noindent\textbf{U}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{kame}{\underline{ka}mE}{v.}{understand}{S}
\word{tslam}{\texttslig lam}{v.}{understand}{S}
\word{'ong}{PoN}{v.}{unfold}{S}
\word{tsawlapxangrr}{\texttslig awlap'aN\textsyllabic{r}}{n.}{unidelta tree}{SG} (for short, {\bf tsawlapx})
\word{silwanin}{silwanin}{}{unknown Cameronian word}{C}
\word{s\`ik}{sIk}{v.}{unquote}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{V}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{kangay}{kaNaj}{adj.}{valid}{S}
\word{\`il\"a}{Il\ae] or [\underline{I}l\ae}{prep.}{via}{S}
\word{nantang}{\underline{nan}taN}{n.}{viperwolf {Caniferratus costatus}, striped armored wolf (Pandoran fauna)}{S}
\word{\"aie}{\ae iE}{n.}{vision}{S}
\word{mokri}{mokRi}{}{voice}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{W}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{pey}{pEj}{v.}{wait}{S}
\word{t\`iran}{tIRan}{v.}{walk}{S}
\word{new}{nEw}{v.}{want}{F} (changed from $_S$ {\bf neu})
\word{tsam}{\texttslig am}{n.}{war}{S}
\cw{tsampongu}{\underline{\texttslig am}poNu}{n.}{war party}{tsam}{war}{pongu}{party}{S}
\word{eyaye}{EjajE}{n.}{warbonnet (Pandoran flora)}{SG}
\cw{tsamsiyu}{\underline{\texttslig am}siju}{n.}{warrior}{tsam si}{make war}{--yu}{agent suffix}{S}
\word{yur}{juR}{v.}{wash}{S}
\word{nari si}{naRi si}{v.}{watch out}{S}
\derive{*pay}{paj}{n.}{water}{paywll}{water plant}{txumpaywll}{poison water plant}{T}
\word{paywll}{pajw\textsyllabic{l}:}{n.}{water plant}{SG}
\word{fya'o}{fjaPo}{n.}{way}{S}
\cw{moe}{moE}{pn.}{we}{me+}{dual}{oe}{I}{S} $_T$
\word{ayoe}{ayoE}{pn.}{we (exclusive)}{S}
\word{ayoeng}{ay\underline{oEN}}{pn.}{we (inlcusive)}{S,MS}
\word{oeng}{oEN}{pn. dual inclusive}{we two}{S}
\cw{moe}{moE}{pn.}{we two (exclusive)}{me+}{dual}{oe}{I}{S} $_T$
\word{meyp}{mEjp}{adj.}{weak}{S}
\word{t\"aftxu}{t\ae ft'u}{v.}{weave}{S}
\cw{t\"aftxuyu}{t\ae\underline{ft'u}ju}{n.}{weaver}{t\"aftxu}{weave}{--yu}{agent suffix}{S}
\word{n\`iltsan}{nIl\texttslig an}{adv.}{well}{S}
\word{Vitraya Ramunong}{vitraja RamunoN}{prop.b}{well of souls}{C}
\word{fpom}{fpom}{}{well-being}{S}
\word{pe+}{pE}{affix inter. marker}{what}{W}
\word{kempe}{kEmpE}{pn.}{what (action)}{S}
\lenite{pehem}{pEhEm}{adv.}{what (action)}{kempe}{S}
\cw{'upe}{P\underline{u}pE}{n.}{what (thing)}{*'u}{thing}{pe}{inter. marker}{S}
\lenite{peu}{pE\underline{u}}{pn.}{what (thing)}{'upe}{S}
\word{krrpe}{k\textsyllabic{r}pE}{adv.}{when}{S}
\lenite{pehrr}{pEh\textsyllabic{r}}{adv.}{when}{krrpe}{S}
\word{peseng}{pEsEN}{adv.}{where}{S}
\word{tsengpe}{\texttslig ENpE}{adv.}{where}{S}
\word{a}{a}{pn.}{which}{S}
\cww{tengkrr}{tEN\underline{k\textsyllabic{r}}}{conj.}{while}{krr}{time}{S}
\lenite{pesu}{pEsu}{pn.}{who}{tupe}{S}
\word{tupe}{\underline{tu}pE}{pn.}{who}{S}
\lenite{lumpe}{lumpE}{adv.}{why}{pelun}{S}
\word{pelun}{pElun}{adv.}{why}{S}
\word{hufwe}{hufwE}{n.}{wind}{S}
\word{tsyal}{\texttslig jal}{n.}{wing}{S}
\cw{txantslusam}{\underline{t'an}\texttslig lusam}{adj.}{wise}{txan}{much}{tsl$<$us$>$am}{understand}{S}
\word{fa}{fa}{prep.}{with}{S}
\word{hu}{hu}{adp.}{with (accompaniment)}{S}
\word{n\`iftxavang}{nI\underline{ft'a}vaN}{adv.}{with all heart}{S}
\word{Mo'at}{\underline{mo}Pat}{prop.n.}{woman's name}{C}
\word{Neyt\`iri}{nEjtIRi}{prop.n.}{woman's name}{C}
\word{Ninat}{nInat}{prop.n.}{woman's name}{C}
\word{l\`i'u}{\underline{lI}Pu}{n.}{word}{S,MS}
\derives{*p\"anu}{p\ae nu}{n.}{word}{p\"anut\`ing}{promise}{D}
\derives{*ler}{lER}{}{work?}{lertu}{colleague}{T}
\word{kifkey}{kifkEj}{n.}{world (solid)}{F}
\word{Eywa}{\underline{Ej}wa}{prop.n.}{world spirit}{C} \textbf{\underline{Ey}wa \underline{nga}hu} intj. {\it good-bye, ... be with you} $_S$
\word{ngawng}{NawN}{n.}{worm (Pandoran fauna)}{S}
\word{'e'al}{P\underline{E}Pal}{}{worst}{S}
\word{pxan}{p'an}{adj.}{worthy}{S}
\word{tewti}{tEwti}{intj.}{wow}{FE}
\end{hangparas} }\end{multicols}

\noindent\textbf{Y}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{aynga}{ayNa}{pn.}{y'all}{S}
\word{z\`is\`it}{\underline{zI}sIt}{n.}{year}{S}
\word{rim}{Rim}{adj.}{yellow}{S}
\word{srane}{sRanE}{}{yes}{S} %POS?
\word{srak(e)}{\underline{sRak}(E)}{part. for}{yes or no questions}{S}
\word{nga}{Na}{pn.}{you}{S}
\word{nga-}{\underline{nga}-}{pn.}{you}{MS}
\word{ngenga}{NENa}{pn.}{you (honorific form)}{S}
\word{aynga}{ayNa}{pn.}{you all}{S}\end{hangparas} }\end{multicols}
\end{document}  