\documentclass[11pt]{article}
\usepackage[left=4cm,top=3cm,right=3cm,left=3cm]{geometry} 
\geometry{a4paper}     
\usepackage{fullpage}
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{tipa}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{epstopdf, ulem}
\usepackage{setspace}
\usepackage{palatino, url, multicol}
\usepackage{hanging, pdfpages, supertabular}
\usepackage[swedish]{babel} \usepackage[utf8]{inputenc} \usepackage[T1]{fontenc}
%\usepackage{fwlw}
\newcommand{\word}[5]{\par\textbf{#1}: [\textipa{#2}] $_{#5}$ #3 \textit{#4}}
\newcommand{\lenite}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#6}$ #3 \textit{#4} allomorph of \textbf{#5}}
\newcommand{\marker}[4]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3}
\newcommand{\cw}[9]{\par\textbf{#1}: [\textipa{#2}] $_{#9}$ #3 \textit{#4} (c.w. from \textbf{#5} \textit{#6} and \textbf{#7} \textit{#8})}
\newcommand{\cww}[7]{\par\textbf{#1}: [\textipa{#2}] $_{#7}$ #3 \textit{#4} (c.w. from \textbf{#5} \textit{#6})}
\newcommand{\loan}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#6}$ #3 \textit{#4} (loaned from Eng. \textit{#5})}
\newcommand{\note}[8]{\par\textbf{#1}: [\textipa{#2}] $_{#8}$ #3 \textit{#4} (#5 \textbf{#6} \textit{#7})}
\newcommand{\derive}[9]{\par\textbf{#1}: [\textipa{#2}] $_{#9}$ #3 \textit{#4} derived from \textbf{#5} \textit{#6} and \textbf{#7} \textit{#8}}
\newcommand{\derives}[7]{\par\textbf{#1}: [\textipa{#2}] $_{#7}$ #3 \textit{#4} derived from \textbf{#5} \textit{#6}}
\newcommand{\deriveall}[8]{\par\textbf{#1}: [\textipa{#2}] $_{#8}$ #3 contraction of \textbf{#4} \textit{#5}, derived from \textbf{#6} \textit{#7}}
\newcommand{\infix}[5]{\par\textbf{#1}: [\textipa{#2}] $_{#5}$ #3 verbal infix in position #4}
\newcommand{\affect}[4]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3 affect verbal infix in position 3}
\newcommand{\affix}[5]{\par\textbf{#1}: [\textipa{#2}] $_{#5}$ #3 affix for #4}
\newcommand{\alloffix}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#6}$ #4 suffix for nouns ending in a #5 (allomorph of \textbf{#3})}
\newcommand{\alloffixx}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#6}$ #4 suffix for #5 (allomorph of \textbf{#3})}
\newcommand{\derivingaffix}[4]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3 deriving affix}
\newcommand{\infixcw}[7]{\par\textbf{#1}: [\textipa{#2}] $_{#7}$ #3 verbal infix in position #4, derived from \textbf{#5} + \textbf{#6}}
\newcommand{\infixcww}[8]{\par\textbf{#1}: [\textipa{#2}] $_{#8}$ #3 verbal infix in position #4, derived from \textbf{#5} + \textbf{#6} or \textbf{#7}}
\newcommand{\infixcwww}[9]{\par\textbf{#1}: [\textipa{#2}] $_{#9}$ #3 verbal infix in position #4, derived from \textbf{#5} + \textbf{#6} or \textbf{#7}, allomorph of \textbf{#8}}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\newenvironment{packed_enum}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}
\title{{\sc Na'vi - English Dictionary} $v.$ 9.15}
\author{Compiled by \href{mailto:richard@learnnavi.org}{Taronyu (Richard Littauer)}\\ \href{www.LearnNavi.org}{www.LearnNavi.org}}
\date{Last updated: February 14$^{th}$, 2010}
\begin{document}
\newpage
\singlespacing
\maketitle
Having been disappointed by the paucity of collated and collected data, I have worked all of the available sources I could find into one dictionary here. Thanks are certainly to be given to the following people for their monumental efforts: Dr. Frommer, Karyu Amawey, all the people in the change log, and everyone on IRC, Skype, and the \href{http://forum.learnnavi.org}{Learn Na'vi} site.
Where appropriate, extra notes have been included for the entries below. Compound non-inflected words and loan words from English have been noted. I have on several occasions derived non-attested root forms: these are marked by a *, and should list sources. The traditional Na'vi transcription in use by Frommer has been maintained throughout: the IPA transcription is given as well. Some, if not many, parts of speech may be erroneous - many have been excluded entirely. Where suitable, lenited words have been included. Some words found elsewhere are not included here, as all affixes and infixes have been avoided, unless their inclusion is deemed particularly useful.  Forms that are inflected have been included if their uninflected form is not attested.\par \textbf{Bold} indicates Na'vi word. {\sl Italics} indicate English definition. \uline{Underlining} indicates where Frommer has given us stress information for polysyllabic words: I have included this in the IPA transcriptions instead of using normal IPA \textprimstress , due to syllable ambiguity. Plain text indicates grammatical categorisation: parentheses indicate further notes. $<$x$>$ represents an infix, where {\it t}$<$1$>$$<$2$>${\it ar}$<$3$>${\it on} are the positions: I have included periods in the IPA to mark where infixes are to be placed, as in {\it t.ar.on}. Proper nouns have been capitalised. Knowledge of lenition processes, including vowel reduction, will be assumed. Elision sometimes occurs, but only in song. For more on inflections, see \href{http://www.talknavi.com/taronyu/Inflections.pdf}{this document.}\\ \\
{\bf Abbreviations:}
\begin{multicols}{2}{\noindent
-- = morpheme boundary\\
+ = leniting morpheme boundary\\
= = eliding boundary\\
$< >$ = infix morpheme\\
` = ejective [in IPA]\\
adj. = adjective\\
adp. = adpositional affix\\
adv. = adverb\\
conj. = conjunction\\
c.w. = compound word\\
dem. = demonstrative\\
intj. = interjection\\
inter. = interrogative\\
lit. = literally\\
n. = noun\\
num. = number\\
part. = particle\\
prep. = preposition\\
pn. = pronoun\\
prop.n. = proper noun\\
v. = verb\\
$\phi$ = illegal phonotactics}
\end{multicols}
For those of you wishing for an English - Na'vi dictionary, use \href{http://content.learnnavi.org/taronyu/NaviEnglish.pdf}{this document}, or use the find function, normally Ctrl + F. 
\section*{Sources}
I have sourced every word as best as I could. This sourcing is placed in a subscript after the IPA transcription, and tells where I, or the source from which I received a word, initially got their word from Frommer. In order of validity, from most to least:\begin{packed_enum}
\item F = Frommer himself.
\item FE = Frommer's personal emails
\item MS = the movie script
\item S = the list at the end of the Survival Guide
\item W = Wikipedia (which has received Frommer's blessing)
\item SG = the Survival Guide main text (flora and songs)
\item M = transcriptions or scripts from the Avatar movie
\item C = Cameron himself (from the 2007 script)
\item T = Taronyu
\item D  = words not derived by Taronyu.
\item G = the Avatar games (which have illegal names)
\item SG = the Survival Guide main text (everything else)
\end{packed_enum}
Frommerian words have the highest worth. The Activists Survival Guide contains a list at the end of Na'vi words, which is the main source of all present dictionaries. Words not in this list but still within the Guide often have illegal phonotactic entries. I have attached the $\phi$ character to each such word, and put them in their own section in the appendix. There are also spelling mistakes: {\it Omaticaya} and {\it skx'awng} are obvious. However, the plant names look distinctly Frommerian: they have the \`i in some names, and exhibit compounding, unlike the rest of the words in the SG text. Somewhat humorously, I've noticed that all but one word from the Guide that doesn't concern Pandoran flora have illegal combinations (the word is {\bf u'imi} {\it streamers}) . I have therefore left this word, and all botanical terms, in the dictionary.
Words from the game are listed \href{http://forum.learnnavi.org/index.php?topic=412.0}{here}. Some words have been derived from the Movie alone. \href{http://en.wikipedia.org/wiki/Na\%27vi_language}{Wikipedia}, as a general grammatical deriving source, has been used for most grammatical items included here. I myself have derived several words: I have marked those here. Trust them at your own risk. Many notes or derivations that I have made with some guesswork are noted with a $_T$. Where MS is given, assume either same stress patterns in S, or MS stress, unless I have created a new entry for inflected words. For more on proper sources, and their history, go \href{http://forum.learnnavi.org/index.php?topic=727.0}{here}.\\
I have edited this dictionary too many times not to assume that it will have errors. {\bf Please}, if you note any, let me know \href{http://forum.learnnavi.org/index.php?topic=341.0}{here}, and I will change them. As for version number, I'll change the first digit when there is a big change. I'll update .0x0 when I make a significant, but not large change. Finally, I'll edit the .00x number when I've made tiny edits. Refer to the Change Log for explanation. Version 2 sourced every word. Version 3 involved adding words from the Guide's main text. Version 4 took illegal words and put them in a separate appendix. Version 5 added an appendix for inflectional endings, with examples, as well as a change log. Version 6 added all original Cameronian words. Version 7 is {\it ta frapo s\`i ne frapo}, and has a lot of new Frommerian words.
\newpage
\noindent{\bf \textipa{\large{P}}}\nopagebreak\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{'ampi}{Pampi}{}{touch}{S}
\word{'angts\`ik\"a}{PaN\texttslig Ik\ae}{n.}{hammerhead (Pandora fauna)}{G}
\word{'aw}{Paw}{num.}{one}{S}
\word{'aw--}{Paw}{third person neutral pronoun}{one}{T}
\word{'awkx}{Pawk'}{n.}{cliff}{S}
\word{'awpo}{Pawpo}{pn.}{one individual}{S} $_{T}$
\word{'awsiteng}{Pawsi\textprimstress tEN}{adv.}{together}{S}
\word{'awve}{PawvE}{adj.}{first}{S} %note n\`iawve
\word{'e'al}{P\uline{E}Pal}{}{worst}{S}
\word{'efu}{P.Ef.u}{v.}{feel}{F}
\word{'eko}{P.\uline{E}k.o}{v.}{attack}{S}
\word{'ekong}{P\uline{E}koN}{n.}{beat (rhythmic)}{S}
\word{'engeng}{P\uline{E}NEN}{n.}{level, floor $_{T}$}{S}
\word{'eveng}{P\uline{E}vEN}{n.}{child}{S}
\note{'evi}{P\uline{E}vi}{n.}{kid}{affectionate form of}{'eveng}{child}{S}
\word{'ewan}{PEwan}{adj.}{young}{F}
\word{'eylan}{PEjlan}{n.}{friend}{F}
\word{'eyng}{P.EjN}{v.}{answer}{F}
\word{'\`i'awn}{P.I\uline{P.awn}}{v.}{remain, stay}{S, MS}
\word{'\`iheyu}{PIhEju}{n.}{spiral}{SG} %from the songs
\loan{'\`ingl\`is\`i}{P\uline{IN}lIsI}{n.}{English language}{English}{S}
\word{'it}{P\uline{i}t\textcorner}{n.}{bit, a small amount}{S}
\word{'itan}{Pitan}{n.}{son}{S}
\word{'ite}{P\uline{i}tE}{n.}{daughter}{S}
\word{'ok}{Pok\textcorner}{n.}{remembrance}{S}
\word{'ong}{P.oN}{v.}{unfold, blossom}{S}
\derive{*'u}{Pu}{n.}{thing}{f\`i'u}{this thing}{tsa'u}{that thing}{T}
\cw{'upe}{P\uline{u}pE}{inter.}{what (thing)}{*'u}{thing}{pe}{inter. marker}{S}
\word{'upxare}{Pup'aRE}{n.}{message}{S}
\end{hangparas} }\end{multicols}
\noindent \textbf{A}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{a}{a}{pn.}{which, that}{S}
\marker{--a--}{a}{adjectival attributive marker}{F}
\word{alaks\`i}{al\uline{ak}sI}{adj.}{ready}{S}
\word{Amhul}{amhul}{prop.n.}{child's name}{S}
\word{Anurai}{anuRai}{prop.n.}{clan name}{G}
\word{al\`im}{alIm}{}{far, far away, at a distance}{?}
\infix{$<$am$>$}{am}{past tense}{2}{W}
\marker{--an}{an}{masculine suffix for certain nouns}{W}
\word{apxa}{ap'a}{adj.}{large}{S}
\infixcw{$<$arm$>$}{aRm}{imperfective past tense}{2}{am}{er}{W}
\word{atan}{atan}{}{light}{S}
\derive{Ateyitan}{atEjitan}{prop.n.}{son of Ateyo}{Ateyo}{man's name}{'itan}{son}{C}
\derives{*atoki}{atoki}{n.}{sacred tree (Pandoran flora)}{atokirina'}{seeds of the great tree}{T}
\cww{atokirina'}{atokiRi\uline{na}P\textcorner}{n.}{seeds of the great tree}{rina'}{seed}{C,S}
\word{atxkxe}{at'k'E}{n.}{land}{S}
\word{au}{au}{n.}{drum (made of skin)}{S}
\word{aungia}{auNia}{n.}{sign, omen}{S}
\word{awaiei}{awaiEi}{n.}{banshee of paradise (Pandoran flora)}{SG}
\derives{awnga}{awNa}{pn.}{we}{ayoeng}{we}{F}
\affix{ay+}{aj}{plural}{nouns}{F}
\infix{$<$ay$>$}{aj}{future tense}{2}{W}
\word{ayfo}{ayfo}{pn.}{they}{S}
\word{aynga}{ayNa}{pn.}{you all}{S}
\word{ayoe}{aywE}{pn.}{we (exclusive)}{S}
\word{ayoeng}{ay\uline{wEN}}{pn.}{we (inclusive)}{S,MS}
\end{hangparas} }\end{multicols}
\noindent \textbf{\"A}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\alloffix{--\"a}{\ae}{y\"a, ey\"a}{genitive}{consonant}{W}
\word{\"aie}{\ae iE}{n.}{vision}{S}
\word{\"Ans\`it}{\uline{\ae n}sIt\textcorner}{prop.n.}{child's name}{S}
\affect{$<$\"ang$>$}{\ae N}{pejorative}{W}
\infix{$<$\"ap$>$}{\ae p}{reflexive}{1}{FE}
\word{\"ao}{\ae o}{prep.}{under}{F}
\word{\"atx\"ale}{\ae \uline{t'\ae}lE}{n.}{request}{S}
\word{\"atx\"ale si}{\ae \uline{t'\ae}lE s.i}{v.}{request}{MS}
\end{hangparas} }\end{multicols}
\noindent\textbf{E}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\marker{--e}{E}{feminine suffix for certain nouns}{F}
\word{ean}{\uline{E}an}{adj.}{blue}{S}
\affect{$<$ei$>$}{Ei}{laudative}{F}
\word{eltu}{\uline{El}tu}{n.}{brain}{S}
\word{eltu si}{\uline{El}tu s.i}{v.}{listen, pay attention}{S}
\cw{eltungawng}{\uline{El}tuNawN}{n.}{brainworm (Pandoran fauna)}{eltu}{brain}{ngawng}{worm}{S} $_T$
\derive{*(')em}{(P)Em}{prep.}{above}{ta'em}{from above}{emza'u}{pass}{T}
\cw{emza'u}{Emz.\uline{a}P.u}{v.}{pass (a test)}{em}{over}{za'u}{come}{S,MS}
\word{eo}{\uline{E}o}{prep.}{before, in front of}{S}
\infix{$<$er$>$}{ER}{Imperfective aspect}{2}{D}
\alloffixx{=ey\"a}{Ej\ae}{\"a, y\"a}{genitive}{pronouns}{W}
\word{eyawr}{ejawR}{adj.}{correct}{FE}
\word{eyaye}{EjajE}{n.}{warbonnet (Pandoran flora)}{SG}
\word{eyk}{.Ejk\textcorner}{v.}{lead}{S}
\word{eyktan}{\uline{Ejk}tan}{n.}{leader}{S}
\word{Eytukan}{\uline{Ej}tukan}{prop.n.}{man's name}{C,MS}
\word{Eywa}{\uline{Ej}wa}{prop.n.}{world spirit, guiding force and deity (equivalent to Gaia)}{C} \textbf{\uline{Ey}wa \uline{nga}hu} intj. {\it good-bye, ... be with you} $_S$
\cw{Eywa'eveng}{EjwaPEvEN}{n.}{Pandora}{Eywa}{Gaia}{'eveng}{child}{F}
\end{hangparas} }\end{multicols}
\noindent\textbf{F}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{fa}{fa}{prep.}{with, by means of}{S}
\word{fahew}{fa\uline{hEw}}{n.}{smell}{F}
\word{faketuan}{fakEtuan}{n.}{alien, non-Na'vi}{?} (might be poor transcription of {\bf fa} + {\bf ketuan}) $_{T}$
\cw{fay+}{faj}{adp.}{these}{f\`i}{this}{ay+}{plural}{D}
\cw{fayvrrtep}{fajv\textsyllabic{r}tEp}{n.}{these demons}{fay}{this plural}{vrrtep}{demon}{F} $_T$
\derive{f\`i=}{fI}{adp.}{this}{f\`ipo}{this one}{f\`i'u}{this thing}{F}
\word{f\`ifya}{fIfja}{dem.}{this, this way, like this}{S} %fja'o
\word{f\`ikem}{fIkEm}{dem.}{this (action)}{S}
\word{f\`ipo}{\uline{fI}po}{dem.}{this one (person or thing)}{S}
\cw{f\`itxan}{fIt'an}{adv.}{so}{f\`i}{this}{txan}{much}{F}
\word{f\`itseng(e)}{fI\uline{\texttslig EN}(E)}{dem.}{this, here, this place}{S,MS}
\word{f\`i'u}{fIPu}{dem.}{this, this thing}{S}
\word{fkarut}{\uline{fka}Rut\textcorner}{}{peel}{S}
\word{fkeu}{fkEu}{adj.}{mighty}{S} (might be {\bf fkew}) $_D$
\word{fko}{fko}{}{one}{S} (might be from {\bf po}) $_{T}$ 
\word{fl\"a}{fl.\ae}{v.}{succeed}{F}
\word{fmawn}{fmawn}{n.}{news, something to report}{S}
\word{fmetok}{fm.\uline{E}t.ok\textcorner}{v.}{test}{S}
\word{fmi}{fm.i}{v.}{try}{S}
\word{fngap}{fNap\textcorner}{n.}{metal}{F}
\word{fngapsutxwll}{fNapsut'w\textsyllabic{l}}{n.}{metal-following plant (Pandoran flora)}{SG}
\word{fnu}{fn.u}{v.}{quiet, be quiet}{S}
\derive{fo}{fo}{pn.}{they}{ay+}{deleted plural leniting morpheme}{po}{he/she}{S} $_T$ %make normal text?
\word{fpak}{fp.ak\textcorner}{v.}{hold off, suspend action}{S}
\word{fpe'}{fp.EP\textcorner}{v.}{send}{S}
\word{fpeio}{fpE\uline{i}o}{}{challenge (ceremonial)}{S}
\word{fpi}{fpi}{prep.}{for the sake of}{S}
\word{fp\`il}{fp.Il}{v.}{think}{FE}
\word{fpom}{fpom}{}{well-being, peace}{S}
\word{fpxafaw}{fp'afaw}{n.}{medusa (Pandoran fauna)}{S}
\word{fpx\"ak\`im}{fp'.\ae k.Im}{v.}{enter}{S}
\derive{*fra--}{fRa}{adp.}{every}{frapo}{everyone}{fra'u}{everything $_D$}{T}
\cww{fra'u}{fRaPu}{n.}{everything}{'u}{thing}{S}
\cww{frakrr}{fRak\textsyllabic{r}}{adv.}{always}{krr}{time}{FE}
\cww{frapo}{\uline{fRa}po}{n.}{everyone}{po}{he/she}{S,MS}
\word{frrfen}{\uline{f.\textsyllabic{r}}f.en}{v.}{visit}{FE}
\word{frrtu}{\uline{f\textsyllabic{r}}tu}{n.}{guest}{FE}
\word{ftang}{ft.aN}{v.}{stop}{S}
\word{fte}{ftE}{conj.}{so that}{S}
\cw{fteke}{ftEkE}{conj.}{so that not, lest}{fte}{so that}{ke}{not}{S}
\word{ftia}{ft.i.\uline{a}}{v.}{study}{F}
\word{ftx\`i}{ft'I}{n.}{tongue}{FE}
\word{ftxoz\"a}{ft'oz\ae}{n.}{holiday}{FE}
\word{ftu+}{ftu}{prep.}{from (direction)}{S}
\word{ftue}{ftuE}{adj.}{easy}{S}
\derives{*ftxavang}{ft'avaN}{adj.}{passionate}{n\`iftxavang}{passionately}{T}
\word{ftxey}{ft'.Ej}{v.}{choose}{S}
\word{fu}{fu}{conj.}{or (+negative{\bf ke}: nor)}{S,FE}
\cw{futa}{futa}{conj.}{that (subordinate clause marker)}{f\`i'u-t}{this accusative}{a}{that}{FE}
\cw{fwa}{fwa}{conj.}{that which}{f\`i'u}{this}{a}{relative clause marker}{F}
\derive{*fya--}{fja}{n.}{path, direction}{fya'o}{path}{fyape}{how}{T}
\word{fya'o}{fjaPo}{n.}{path, way}{S}
\lenite{fyape}{fjapE}{inter.}{how}{pefya}{S}
\cw{fyaw\`intxu}{fjaw.In\uline{t'.u}}{v.}{guide}{*fya}{path}{w\`intxu}{show}{S} $_T$
\end{hangparas} }\end{multicols}
\noindent\textbf{H}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{ha}{ha}{conj.}{so, in that case}{S}
\word{hahaw}{h.ah.aw}{v.}{sleep}{F}
\word{hapx\`i}{ha\uline{p'I}}{}{part}{S,MS}
\word{hasey}{hasEj}{adj.}{done, finished}{S}
\word{hawnu}{h.awn.u}{v.}{protect, shelter}{S}
\word{h\`i'i}{\uline{hI}Pi}{adj.}{small}{S}
\word{hiy\`ik}{\uline{hi}jIk\textcorner}{adj.}{funny, strange}{S}
\word{hol}{hol}{adj.}{few}{FE}
\word{holpxay}{holp'aj}{n.}{number}{FE}
\word{hrrap}{h\textsyllabic{r}ap}{n.}{danger}{S}
\word{hu}{hu}{prep.}{with (accompaniment)}{S}
\word{hufwe}{hufwE}{n.}{wind}{S}
\word{hum}{h.um}{v.}{leave, depart}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{I}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{i'en}{iPEn}{n.}{stringed instrument}{S}
\derive{*ik}{ik\textcorner}{n.}{heaven, sky, mountain}{ik-ni-maya}{stairway to heaven}{ikran}{walks the sky}{D}
\word{Iknimaya}{\uline{ik}ni\uline{ma}ja}{prop.n.}{Stairway to Heaven, the rite of passage for young Na'vi}{C,S}
\word{ikran}{\uline{ik}Ran}{n.}{banshee}{C} ({\it mountain banshee}) $_{SG}$
\word{ikranay}{ikRanaj}{n.}{forest banshee}{SG} (dubious: --ay would make {\it banshees}) $_T$
\word{ioang}{i\uline{o}aN}{n.}{animal, beast}{S}
\word{irayo}{iRajo}{n., intj.}{thank you}{F} (also {\bf ireiyo}) $_{MS}$
\infixcw{$<$ilv$>$}{ilv}{present perfective subjunctive}{2}{iv}{ol}{FE}
\infixcww{$<$imv$>$}{imv}{past subjunctive}{2}{iv}{am}{\`im}{FE}
\infixcw{$<$irv$>$}{iRv}{present imperfective subjunctive}{2}{iv}{er}{FE}
\alloffix{--it}{it}{t, ti}{accusative}{consonant}{W}
\infix{$<$iv$>$}{iv}{subjunctive mood}{2}{W}
\infixcwww{$<$iyev$>$}{ijEv}{future subjunctive}{2}{iv}{\`iy}{ay}{$<$\`iyev$>$}{FE}
\end{hangparas} }\end{multicols}
\noindent\textbf{\`I}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\alloffix{--\`il}{Il}{l}{ergative case}{consonant}{W}
\word{\`il\"a}{Il\ae] or [\uline{I}l\ae}{prep.}{by, via, following}{S}
\derives{*\`im}{Im}{prep.}{behind, back}{ne'\`im}{back}{T}
\infix{$<$\`im$>$}{Im}{recent past}{2}{W}
\alloffix{--\`iri}{IRi}{ri}{topical}{consonant}{W}
\infixcw{$<$\`irm$>$}{IRm}{imperfective recent past tense}{2}{\`im}{er}{W}
\infix{$<$\`iy$>$}{Ij}{immediate future tense}{2}{W}
\infixcwww{$<$\`iyev$>$}{IjEv}{future subjunctive}{2}{iv}{\`iy}{ay}{$<$iyev$>$}{FE}
\end{hangparas} }\end{multicols}
\newpage
\noindent\textbf{K}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{ka}{ka}{prep.}{across}{S}
\word{k\"a}{k.\ae}{v.}{go}{S}
\word{kali'weya}{kaliPwEja}{n.}{species of poisonous arachnid (Pandoran fauna)}{SG}
\word{kalin}{kalin}{adj.}{sweet}{FE}
\word{kaltx\`i}{kalt'I}{intj.}{hello}{S}
\cw{k\"amakto}{k\ae\uline{m.ak}t.o}{v.}{ride out}{k\"a}{go}{makto}{ride $_T$}{S}
\word{kame}{\uline{k.a}m.E}{v.}{see, see into, understand, know}{S}
\word{kan}{k.an}{v.}{aim}{FE}
\word{kangay}{kaNaj}{adj.}{valid}{S}
\derives{*kar}{k.aR}{v.}{teach}{karyu}{teacher}{T}
\word{karyu}{kaRju}{n.}{teacher}{S}
\word{kato}{\uline{ka}to}{n.}{rhythm}{S}
\deriveall{*kaw--}{kaw}{adp.}{ke aw}{not one}{kawkrr}{never}{T}
\cww{kawkrr}{\uline{kaw}k\textsyllabic{r}}{adv.}{never}{krr}{time}{S}
\cw{kawng}{kawN}{adj.}{bad, evil}{*kaw--}{not}{--ng}{inclusive}{S} $_T$
\cww{kawtu}{kawtu}{pn.}{no-one}{tu}{person}{S}
\word{ke}{kE}{adv., adj.}{not}{S}
\word{ke--}{kE}{adp.}{no, not}{T}
\cw{ke'u}{\uline{kE}Pu}{pn.}{nothing}{ke}{not}{'u}{thing}{S}
\word{kehe}{\uline{kE}hE}{part.}{no}{S}
\word{kelku}{\uline{kEl}ku}{n.}{home}{S}
\word{kelku si}{kElku s.i}{v.}{live, dwell}{S}
\word{Kelutral}{\uline{kE}lutRal}{prop.n.}{Hometree (Pandoran flora)}{S,MS}
\derive{*kem}{kEm}{}{action}{f\`ikem}{this action}{kempe}{what action}{T}
\word{kempe}{kEmpE}{inter.}{what (action)}{S}
\word{kenong}{\uline{k.E}n.oN}{v.}{model, represent, exemplify}{S}
\cw{kerusey}{\uline{kE}RusEj}{adj.}{dead}{ke}{not}{r$<$us$>$ey}{living}{S} $_D$
\word{ketuwong}{\uline{kE}tuwoN}{n.}{alien}{S}
\word{kewong}{\uline{kE}woN}{adj.}{alien}{S}
\word{keye'ung}{kE\uline{jE}PuN}{n.}{insanity}{S}
\word{k\`i'ong}{kIPoN}{n.}{kind of fruit or vegetable (Pandoran flora)}{S}
\word{kifkey}{kifkEj}{n.}{world (solid)}{F}
\word{kilvan}{kilvan}{n.}{river}{F}
\word{k\`im}{k.Im}{v.}{spin}{S}
\word{kin}{k.in}{v.}{need}{S}
\word{kin\"a}{\uline{ki}n\ae}{num.}{seven}{S}
\word{kinam}{kinam}{n.}{leg}{S}
\word{kinamtil}{ki\uline{nam}til}{n.}{knee}{S}
\word{k\`ing}{kIN}{}{thread}{S}
\word{kip}{kip\textcorner}{prep.}{among}{S}
\word{k\`ite'e}{kI\uline{tE}PE}{}{service}{S}
\cw{k$<$\`iyev$>$ame}{kIjE\uline{va}m.E}{intj.}{good-bye, see you soon}{kame}{see}{$<$\`iyev$>$}{future subjunctive infix}{S}
\derive{*kll--}{k\textsyllabic{l}}{}{ground}{kllkulat}{dig up}{kllkxem}{stand}{T}
\word{kllfrivo'}{k\textsyllabic{l}fR.i\uline{v.o}P}{v.}{be responsible}{S}
\word{kllkulat}{k\textsyllabic{l}\uline{k.u}l.at}{v.}{dig up}{S}
\word{kllkxem}{k.\textsyllabic{l}k'.Em}{v.}{stand}{S}
\word{kllpx\`iltu}{k\textsyllabic{l}\uline{p'Il}tu}{n.}{territory}{S}
\word{kllte}{k\textsyllabic{l}tE}{n.}{ground}{S}
\word{ko}{ko}{part.}{solicit agreement}{FE}
\word{kop}{kop\textcorner}{part.}{too, also}{F}
\word{koren}{koREn}{n.}{rule}{F}
\word{krr}{k\textsyllabic{r}}{n.}{time}{S}
\word{*krr}{k\textsyllabic{r}}{adv.}{ever, time}{M} (derived from direct transcription)
\word{krrnekx}{k\textsyllabic{r}n.Ek'}{v.}{take/consume time}{S}
\word{krrpe}{k\textsyllabic{r}pE}{inter.}{when}{S}
\loan{kuns\`ip}{\uline{kun}sIp\textcorner}{n.}{gunship}{gunship}{S}
\word{kurakx}{k.uR.ak'}{v.}{drive out}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{Kx}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{kxam}{k'am}{n.}{middle, midpoint}{S}
\cw{kxamtseng}{\uline{k'am}\texttslig EN}{n.}{center-place}{kxam}{center}{tseng}{place}{S}
\word{kxangangang}{\uline{k'aN}aNaN}{intj.}{boom (onomatopoeia)}{S}
\word{kxan\`i}{k'anI}{adj.}{forbidden}{S}
\word{kxawm}{k'awm}{adv.}{perhaps, maybe}{S}
\word{kxener}{\uline{k'E}nER}{n.}{kind of fruit or vegetable (Pandoran flora)}{S}
\word{kxetse}{\uline{k'E}\texttslig E}{n.}{tail}{S}
\word{kxeyey}{k'EjEj}{n.}{error, mistake}{F}
\word{kxll}{k'\textsyllabic{l}}{}{charge}{G}
\end{hangparas} }\end{multicols}
\noindent\textbf{L}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\affix{--l}{l}{ergative case}{nouns ending in a vowel}{W}
\derive{*la--}{la}{adp.}{other, otherness}{lapo}{other one}{aylahe}{others}{T}
\derive{*lay+}{laj}{adp.}{others}{*la}{other}{ay+}{plural}{D}
\word{lahe}{\uline{la}hE}{pn.}{other}{S}
\word{lam}{l.am}{v.}{seem, appear}{S}
\cww{lapo}{\uline{la}po}{pn.}{other one}{po}{he/she}{S}
\word{latem}{\uline{l.a}t.Em}{v.}{change}{S}
\word{law}{law}{adj.}{clear, certain}{S}
\derivingaffix{le--}{lE}{adjective}{F}
\word{lefpom}{lEfpom}{adj.}{happy, peaceful, joyous (not for people)}{FE}
\word{lertu}{lERtu}{n.}{colleague}{S}
\word{letrr}{lEt\textsyllabic{r}}{adj.}{daily}{T,F}
\cw{letsunslu}{lE\texttslig unslu}{adj.}{possible}{tsun}{be able}{slu}{become}{FE}
\word{Li'ona}{liPona}{prop.n}{clan name}{G}
\derive{*l\`i(')}{lI(P)}{}{speak}{l\`i'u}{word}{l\`i'fya}{language}{W}
\cw{l\`i'fya}{lIfja}{n.}{language}{l\`i'u}{word}{fya}{way $_W$}{F}
\word{l\`i'u}{\uline{lI}Pu}{n.}{word}{S,MS}
\word{lok}{lok\textcorner}{prep., v.}{close, close to}{S}
\word{Lompo}{lompo}{prop.n}{man's name}{G}
\word{lonataya}{lonataja}{n.}{medusa (Pandoran Fauna)}{SG}
\word{lonu}{l.on.u}{v.}{release, let go}{S}
\word{lor}{loR}{adj.}{beautiful (non-human)}{F}
\derive{loreyu}{loREju}{n.}{helicoradium spirale, beautiful spiral (Pandoran Flora)}{lor}{beautiful}{\`ireyu}{spiral}{SG} $_T$
\derives{*lrr}{l\textsyllabic{r}}{}{smile, happy}{lrrtok}{smile exists}{T}
\word{lrrtok}{\uline{l\textsyllabic{r}}tok}{n.}{smile}{S}
\word{lrrtok si}{\uline{l\textsyllabic{r}}tok s.i}{v.}{smile}{M}
\word{lu=}{l.u}{v.}{be (predicative copula, existential verb)}{S}
\word{lun}{lun}{n.}{reason}{F}
\lenite{lumpe}{lumpE}{inter.}{why}{pelun}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{M}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{ma}{ma}{part.}{vocative marker}{M}
\word{makto}{m.akt.o}{v.}{ride}{S}
\word{Malimi}{malimi}{prop.n}{a Na'vi name}{G}
\word{mawey}{mawEj}{adj.}{calm}{S}
\affix{me+}{\uline{mE}}{dual}{nouns}{W}
\cw{mefo}{\uline{mE}fo}{pn.}{they, those two}{me+}{dual marker}{po}{he/she}{S}
\cw{menari}{mE\uline{na}Ri}{n.}{eyes}{me+}{dual}{nari}{eye}{S}
\word{meoauniaea}{mEoauniaEa}{}{Frommer's favourite word, currently undefined}{F}
\word{meuia}{mE\uline{u}ia}{}{honour}{S}
\word{meyp}{mEjp\textcorner}{adj.}{weak}{S}
\word{mi}{mi}{adv.}{yet, still}{FE}
\word{mip}{mip\textcorner}{adj.}{new}{FE}
\word{mikyun}{\uline{mik}jun}{n.}{ear}{C,S}
\word{m\`i+}{mI}{prep.}{in}{S}
\word{m\`in}{mIn}{}{turn}{S}
\derives{m\`iso}{mIso}{prep.}{away (position)}{*to}{away}{S}
\word{mllte}{m.\textsyllabic{l}\uline{t.E}}{v.}{agree}{S}
\word{Mo'at}{\uline{mo}Pat\textcorner}{prop.n.}{woman's name}{C}
\derive{Mo'at'ite}{moPatPitE}{prop.n.}{daughter of Mo'at}{Mo'at}{woman}{ite}{daughter}{C}
\cw{moe}{moE}{pn.}{we, we two (exclusive)}{me+}{dual}{oe}{I}{S} $_T$
\word{mokri}{mokRi}{n.}{voice}{S}
\word{mrr}{m\textsyllabic{r}}{num.}{five}{F}
\word{mui\"a}{mui\ae}{adj.}{proper, fair, right}{S}
\word{mun'i}{m.unP.\uline{i}}{v.}{cut}{S}
\word{mune}{\uline{mu}nE}{num.}{two}{S}
\word{munge}{\uline{m.u}N.E}{v.}{take, bring}{S}
\word{mungwrr--}{muNw\textsyllabic{r}}{adp.}{except}{FE}
\word{muntxa}{munt'a}{adj.}{mated}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{N}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{na}{na}{prep.}{like, as}{M,S}
\word{na'r\`ing}{\uline{na}PRIN}{n.}{forest}{S}
\word{Na'vi}{\uline{na}Pvi}{prop.n.}{the People, indigenous Pandoran sentient race}{C,S,MS}
\word{nang}{naN}{part.}{surprise, exclamation}{S}
\word{nantang}{\uline{nan}taN}{n.}{viperwolf {Caniferratus costatus}, striped armored wolf (Pandoran fauna)}{S}
\word{nari}{\uline{na}Ri}{n.}{eye}{C,S}
\word{nari si}{naRi s.i}{v.}{watch out, be careful}{S}
\word{nawm}{nawm}{adj.}{great, noble}{S}
\word{ne}{nE}{prep.}{to, towards}{S}
\cw{ne kllte}{nE \uline{k\textsyllabic{l}}tE}{intj.}{get down}{ne}{to}{kllte}{the ground}{S}
\word{ne'\`im}{nEP\uline{Im}}{prep.}{back (direction)}{S}
\word{nekx}{n.Ek'}{v.}{burn, consume}{S}
\word{Nera'nak}{nERaPnak\textcorner}{prop.n}{a Na'vi name}{G}
\derive{neto}{nEto}{prep.}{away (direction)}{ne}{to}{*to}{away}{S}
\word{new}{n.Ew}{v.}{want}{F} (changed from $_S$ {\bf neu})
\word{Neytiri}{nEjtiRi}{prop.n.}{woman's name}{C}
\word{ni\"a}{n.i.\ae}{v.}{grab}{S}
\word{Ninat}{ninat\textcorner}{prop.n.}{woman's name, good singer}{C}
\derivingaffix{n\`i=}{nI}{adverbial}{W}
\word{n\`i'aw}{nIPaw}{adv.}{only}{S}
\cw{n\`iawnomum}{nIaw\uline{no}mum}{phrase}{as we know}{ayoe $\rightarrow$ aw}{we}{omum}{know}{F}
\word{n\`i'awtu}{nIP\uline{aw}tu}{adv.}{alone}{S}
\word{n\`i'awve}{nIP\uline{aw}vE}{adv.}{first}{S}
\cww{n\`i'it}{nIPit\textcorner}{adv.}{bit, a small amount (abstract)}{'it}{bit, a small amount}{S}
\word{n\`i'ul}{nIPul}{adv.}{more}{S} 
\word{n\`iayoeng}{nIajwEN] or [naj\uline{wEN}}{adv.}{like us}{S} (corrected from {\bf -eg}) $_T$
\word{n\`iftue}{nI\uline{ftu}E}{adv.}{easily}{S}
\word{n\`iftxavang}{nI\uline{ft'a}vaN}{adv.}{passionately, with all heart}{S}
\word{n\`ihawng}{nIhawN}{adv.}{too, excessively}{S}
\word{n\`iltsan}{nIl\texttslig an}{adv.}{well}{S}
\word{n\`imun}{nImun}{adv.}{again}{S}
\word{n\`in}{n.In}{v.}{look, look at}{S}
\word{n\`ingay}{nINaj}{adv.}{truly}{S} %pronounce?
\word{n\`itam}{nItam}{adv.}{enough}{S} %tam
\word{n\`iteng}{nItEN}{adv.}{too, also}{F}
\word{n\`itut}{nItut\textcorner}{adv.}{continually}{S}
\cww{n\`itxan}{nIt'an}{adv.}{much (abstract)}{txan}{much}{S}
\word{n\`iwin}{nIwin}{adv.}{fast}{S}
\word{n\`iwotx}{nIwot'}{adv.}{all (of), in toto, completely}{S}
\word{Nok}{nok\textcorner}{prop.n}{man's name}{G}
\derive{nul--}{nul}{adp.}{comparative adjectival marker}{n\`i'ul}{more}{nulkrr}{longer}{T,D}
\word{nulkrr}{\uline{nu}lk\textsyllabic{r}}{}{longer (time)}{S}
\word{nulnew}{nuln.ew}{v.}{prefer}{F}
\word{nume}{n.um.E}{v.}{learn}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{Ng}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\affix{--ng}{N}{clusivity, contraction of nga}{pronouns}{W}
\word{nga}{Na}{pn.}{you}{S}
\word{ngawng}{NawN}{n.}{worm (Pandoran fauna)}{S}
\word{ngay}{Naj}{adj.}{true}{S}
\word{ngenga}{NENa}{pn.}{you (honorific form)}{S}
\word{ngim}{Nim}{adj.}{long}{F}
\word{ngop}{N.op\textcorner}{v.}{create}{F}
\derives{*ngrr}{ng\textsyllabic{r}}{n.}{root}{tsawlapxangrr}{tall large root}{D}
\word{nguway}{Nuwaj}{n.}{howl}{G}
\end{hangparas} }\end{multicols}
\noindent\textbf{O}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{oe}{\uline{w}E}{pn.}{I, me}{S}
\word{oeng}{wEN}{pn. dual inclusive}{we two}{S}
\word{oeyk}{o\uline{Ejk}}{n.}{cause}{FE}
\word{ohe}{\uline{o}hE}{pn.}{I (deferential or ceremonial form)}{S, MS}
\word{ohe-}{o\uline{hE}-}{pn.}{I}{MS}
\infix{$<$ol$>$}{ol}{perfective aspect}{2}{W}
\word{olo'}{oloP\textcorner}{n.}{clan}{S}
\cw{olo'eyktan}{\uline{o}loP\uline{Ejk}tan}{n.}{clan leader}{olo'}{clan}{eyktan}{leader}{C,S} (male clan leader) $_{SG}$
\word{Omati}{omati}{prop.n}{man's name}{G}
\word{Omatikaya}{omati\uline{ka}ja}{prop.n.}{Blue Flute Clan}{C,S}
\word{omum}{.\uline{o}m.um}{v.}{know}{S}
\word{Onantu}{onantu}{prop.n}{a Na'vi name}{G}
\word{ontu}{\uline{o}ntu}{n.}{nose}{C,S}
\end{hangparas} }\end{multicols}
\newpage
\noindent\textbf{P}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{p\"ahem}{\uline{p.\ae}h.Em}{v.}{arrive}{S}
\marker{pak}{pak\textcorner}{part. for disparagement}{S}
\word{pa'li}{paPli}{n.}{direhorse (Pandoran fauna)}{S}
\word{palulukan}{\uline{pa}lu\uline{lu}kan}{n.}{Thanator}{S}
\word{pam}{pam}{n.}{sound}{S}
\cw{pamtseo}{\uline{pam}\texttslig Eo}{n.}{music}{pam}{sound}{tseo}{art}{S}
\word{pamtseowll}{pam\texttslig Eow\textsyllabic{l}}{n.}{cat ear, music plant (Pandoran flora)}{SG}
\word{p\"angkxo}{p.ANk'.o}{v.}{chat, converse}{F}
\derives{*p\"anu}{p\ae nu}{n.}{oath, word, promise}{p\"anut\`ing}{promise}{D}
\cww{p\"anut\`ing}{\uline{p\ae}nut.IN}{v.}{promise}{t\`ing}{give}{S}
\word{p\"ats\`i}{\uline{p\ae}\texttslig I}{n.}{badge}{S}
\word{pawm}{p.awm}{v.}{ask}{S}
\derive{*pay}{paj}{n.}{water}{paywll}{water plant}{txumpaywll}{poison water plant}{T}
\word{paywll}{pajw\textsyllabic{l}}{n.}{water plant, dapophet (Pandoran flora)}{SG}
\word{pe+}{pE}{affix inter. marker}{what}{W}
\word{pe'un}{p.EP.un}{v.}{decide}{F}
\lenite{pefya}{pEfja}{inter.}{how}{fyape}{S}
\lenite{pehem}{pEhEm}{inter.}{what (action)}{kempe}{S}
\lenite{pehrr}{pEh\textsyllabic{r}}{inter.}{when}{krrpe}{S}
\lenite{pelun}{pElun}{inter.}{why}{lumpe}{S}
\word{peng}{p.EN}{v.}{tell}{S}
\cw{penghrrap}{pENh\textsyllabic{r}ap}{n.}{binary sunshine, fringed lamp (Pandoran flora)}{peng}{tell}{hrrap}{danger}{SG}
\lenite{peseng}{pEsEN}{inter.}{where}{tsengpe}{S}
\lenite{pesu}{pEsu}{inter.}{who}{tupe}{S}
\lenite{peu}{pE\uline{u}}{inter.}{what (thing)}{'upe}{S}
\word{pey}{p.Ej}{v.}{wait}{S}
\word{Peyral}{pEjRal}{prop.n}{name }{C, M}
\derive{pizayu}{\uline{pi}zaju}{n.}{ancestor}{*za}{come}{yu}{noun deriving affix $_T$}{S}
\word{plltxe}{p.\textsyllabic{l}t'.E}{v.}{speak}{S}
\word{po}{po}{pn.}{he, she}{S}
\word{poan}{poan}{pn.}{he}{S}
\word{poe}{pw\uline{E}}{pn.}{she}{S}
\word{pongu}{\uline{po}Nu}{n.}{group, party}{S}
\word{prrnen}{p\textsyllabic{r}nEn}{n.}{infant, baby}{FE}
\cw{prrnesyul}{p\textsyllabic{r}nEsjul}{n.}{bud}{prrnen}{baby}{syulang}{flower}{FE}
\word{prrte'}{p\textsyllabic{r}tEP}{n.}{pleasure}{F}
\word{pukap}{\uline{pu}kap\textcorner}{num.}{six}{FE}
\end{hangparas} }\end{multicols}
\noindent\textbf{Px}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{pxan}{p'an}{adj.}{worthy}{S}
\word{pxasik}{p'asik\textcorner}{intj.}{screw that!}{F}
\word{pxasul}{\uline{p'a}sul}{adj.}{fresh, appealing as food}{S}
\word{pxay}{p'aj}{adj., pn.}{many}{S}
\affix{pxe+}{\uline{p'E}}{trial}{nouns and pronouns}{F}
\word{pxel}{p'El}{conj.}{like, as}{S}
\word{pxey}{p'Ej}{num.}{three}{F}
\word{pxi}{p'i}{adj.}{sharp}{S}
\word{px\`im}{p'Im}{adv.}{often}{F}
\word{pxiut}{p'iut\textcorner}{n.}{razor palm, sharp tree (Pandoran flora)}{SG}
\cww{pxorna'}{p'ornaP\textcorner}{n.}{episoth, exploding seed (Pandoran flora)}{rina'}{seed}{SG} $_T$
\word{pxun}{p'un}{n.}{arm}{S}
\word{pxuntil}{p'until}{n.}{elbow}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{R}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{r\"a'\"a}{R\ae P\ae}{part.}{do not (for negative imperatives)}{S}
\word{Rai'uk}{RaiPuk\textcorner}{prop.n}{name}{G}
\word{ral}{Ral}{n.}{meaning}{S}
\cw{ralpeng}{Ralp.EN}{v.}{interpret}{ral}{meaning}{peng}{tell}{S}
\word{Ralu}{\uline{Ra}lu}{prop.n.}{child's name}{S}
\derive{*ran}{R.an}{v.}{contracted form of walk}{t\`iran}{walk}{ikran}{walks-the-heavens}{D}
\word{rawke}{RawkE}{n.}{alarm cry}{S}
\word{rel}{REl}{n.}{image, picture}{F}
\cw{reltseo}{REl\texttslig Eo}{n.}{visual art}{rel}{image}{tseo}{art}{F}
\cww{reltseotu}{REl\texttslig Eotu}{n.}{artist}{reltseo}{art}{F}
\word{renu}{\uline{RE}nu}{n.}{pattern}{S}
\word{rey}{R.Ej}{v.}{live}{S}
\affix{--ri}{Ri}{topical}{nouns ending with a vowel}{F}
\word{rikx}{R.ik'}{v.}{move, shift position}{S}
\word{rim}{Rim}{adj.}{yellow}{S}
\word{rina'}{RinaP\textcorner}{n.}{seed}{S}
\word{riti}{Riti}{n.}{stingbat (Pandoran fauna)}{S}
\word{rol}{R.ol}{v.}{sing}{S}
\word{Rongloa}{RoNloa}{prop.n}{assumed clan name}{C}
\word{ronsem}{\uline{Ron}sEm}{n.}{mind}{S}
\alloffix{--ru}{Ru}{--ur}{dative}{vowel}{W}
\derives{*rum}{Rum}{n.}{ball}{rumut}{ball tree}{T}
\word{rumut}{Rumut\textcorner}{n.}{ball tree (Pandoran flora)}{SG}
\word{run}{R.un}{v.}{find, discover}{FE}
\word{rutxe}{Rut'E}{part.}{please}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{S}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\derive{*s\"a--}{s\ae}{adp.}{instrumental noun}{s\"anume}{teaching}{s\"aspxin}{disease}{T,F}
\word{sa'nok}{saPnok\textcorner}{n.}{mother}{S} (from {\bf sa'atenuk} $_C$)
\note{sa'nu}{saPnu}{n.}{mum}{affectionate form of}{sa'nok}{mother}{S}
\word{saleu}{sa\uline{l.E}.u}{v.}{go}{MS} (might be from {\bf za'u}) $_T$
\word{san}{san}{part.}{saying, quote}{S}
\word{s\"anume}{s\ae\uline{nu}mE}{n.}{teaching, instruction}{S}
\cww{s\"aspxin}{s\ae sp'in}{n.}{disease}{spxin}{sick}{FE}
\note{sempu}{sEmpu}{n.}{dad}{affectionate form of}{sempul}{father}{S}
\word{sempul}{\uline{sEm}pul}{n.}{father}{S}
\word{set}{sEt\textcorner}{adv.}{now}{S}
\word{sevin}{sEvin}{adj.}{pretty}{S} (only for people) $_F$
\word{seyri}{sEjRi}{n.}{lip}{C,S}
\word{Seyvaro}{sEjvaro}{prop.n}{Na'vi name}{G}
\word{seze}{\uline{sE}zE}{n.}{blue flower (Pandoran flora)}{S}
\derivingaffix{si}{si}{{\it do, make} auxiliary verb, postpositional non-bound verb}{S, F}
\note{s\`i=}{sI}{conj.}{and}{connects two things: for clauses use, can be attached as --s\`i}{ulte}{and}{S}
\word{s\`ik}{sIk\textcorner}{part.}{unquote}{S}
\word{s\`ilpey}{s.Ilp.Ej}{v.}{hope}{F}
\cw{s\`ilronsem}{sIl\uline{Ron}sEm}{adj.}{clever (thing)}{s\`iltsan}{good}{ronsem}{mind}{S}
\word{s\`iltsan}{sIl\texttslig an}{adj.}{good}{S}
\word{Silwanin}{silwanin}{prop.n.}{Neytiri's sister}{C,G}
\derive{*s\`ip}{sIp\textcorner}{n.}{ship}{taws\`ip}{sky-ship}{kuns\`ip}{gunship}{T}
\word{ska'a}{sk.aP.\uline{a}}{v.}{destroy}{S}
\word{skxawng}{sk'awN}{n.}{moron}{S}
\word{sl\"a}{sl\ae}{conj.}{but}{S}
\word{slu}{sl.u}{v.}{become}{S}
\word{sng\"a'i}{sN.\ae P.i}{v.}{begin, start}{S}
\cw{sng\"a'ikrr}{\uline{sN\ae}Pik\textsyllabic{r}}{n.}{beginning, starting time}{sng\"a'i}{begin}{krr}{time}{S}
\word{sngel}{sNEl}{n.}{garbage}{F}
\cw{sngeltseng}{sNEl\texttslig EN}{n.}{garbage place}{sngel}{garbage}{tseng}{place}{F}
\word{snum\`ina}{\uline{snu}mIna}{adj.}{dim (of a person)}{S}
\word{som}{som}{adj.}{hot}{S}
\cw{somt\`ilor}{somtIlor}{n.}{popsicle, hot beauty (Pandoran flora)}{som}{hot}{t\`ilor}{beauty}{SG}
\word{spaw}{sp.aw}{v.}{believe}{F}
\derives{*spxam}{sp'am}{n.}{mushroom}{torukspxam}{octoshroom}{T}
\word{spe'e}{sp.EP.\uline{E}}{v.}{capture}{S}
\cw{spe'etu}{spEP\uline{E}tu}{n.}{captive}{spe'e}{capture}{tu}{person}{S}
\word{spxin}{sp'in}{adj.}{sick}{FE}
\word{srak(e)}{\uline{sRak}(E)}{part. for}{yes or no questions}{S}
\word{srane}{sRanE}{part.}{yes}{S}
\word{sreu}{sR.E.u}{v.}{dance}{S} (might be {\bf srew}) $_D$
\word{srung}{sRuN}{n.}{help, assistance}{S}
\word{stawm}{st.awm}{v.}{hear}{F}
\word{steftxaw}{st.Eft'.aw}{v.}{examine}{S}
\word{stum}{stum}{adv.}{almost}{S}
\word{sutx}{s.ut'}{v.}{track, lock up}{S} % should be trap?
\word{swaw}{swaw}{n.}{moment}{S}
\word{swey}{swEj}{adj.}{best}{F}
\word{swir\"a}{swiR\ae}{n.}{creature}{S}
\word{swizaw}{swizaw}{n.}{arrow}{S}
\word{swok}{swok\textcorner}{adj.}{sacred}{S}
\word{swotu}{swotu}{n.}{sacred place}{S}
\word{swotulu}{swotulu}{n.}{sacred river}{G}
\word{syaw}{sj.aw}{v.}{call}{S}
\word{syulang}{sjulaN}{n.}{flower}{FE}
\end{hangparas} }\end{multicols}
\newpage
\noindent\textbf{T}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\alloffix{--t}{t}{it, ti}{accusative}{vowel}{W}
\word{ta}{ta}{prep.}{from (various uses)}{S}
\word{ta'em}{taPEm}{prep.}{from above}{S}
\word{tafral}{tafRal}{adv.}{therefore, because of that}{FE}
\word{t\"aftxu}{t.\ae ft'.u}{v.}{weave}{S}
\cw{t\"aftxuyu}{t\ae\uline{ft'u}ju}{n.}{weaver}{t\"aftxu}{weave}{--yu}{agent suffix}{S}
\word{takuk}{\uline{t.a}k.uk\textcorner}{v.}{strike}{SG}
\cww{talioang}{\uline{tal}i\uline{o}aN}{n.}{sturmbeest (Pandoran fauna)}{ioang}{animal}{S}
\word{talun(a)}{talun(a)}{conj.}{because, from the reason}{P,F}
\word{tam}{t.am}{v.}{suffice, "do"}{S}
\word{tam tam}{tam tam}{intj.}{calm, be content, there there}{M, T}
\word{tangek}{taNEk\textcorner}{n.}{trunk (of a tree)}{S}
\word{tanh\`i}{tan\uline{hI}}{n.}{star}{S}
\word{taron}{\uline{t.a}R.on}{v.}{hunt}{S}
\cw{taronyu}{\uline{ta}Ronju}{n.}{hunter}{taron}{hunt}{--yu}{agent suffix}{C,S}
\word{Tasun}{tasun}{prop.n}{name}{G}
\cw{tautral}{tautRal}{n.}{sky tree (Pandoran flora)}{taw}{sky}{utral}{tree}{SG} $_T$
\word{taw}{taw}{n.}{sky}{F}
\word{taweyk(a)}{tawEjk(a)}{conj.}{because, based on the cause}{FE}
\cw{Tawkami}{tawkami}{prop.n}{clan name}{taw}{sky}{kame}{see}{G}
\loan{taws\`ip}{tawsIp\textcorner}{n.}{sky-ship}{ship}{F}
\word{tawng}{t.awN}{v.}{duck, dive}{S} %eliminated duck n.
\cww{tawtsngal}{taw\texttslig Nal}{n.}{panopyra, sky cup (Pandoran flora)}{taw}{sky}{SG}
\cw{Tawtute}{\uline{taw}tutE}{prop.n.}{Sky-person, human}{taw}{sky}{tute}{person}{S,MS}
\word{te}{tE(P)}{part.}{used in Cameronian full names}{C}
\word{tei}{tEi}{n.}{plains}{G}
\word{tel}{t.El}{v.}{receive}{F}
\word{telem}{tElEm}{n.}{cord}{S} (could be {\it chord}) $_T$
\derive{*teng--}{tEN}{adp.}{like, same as}{tengfya}{same way as}{tengkrr}{while}{T}
\cw{tengfya}{tENfja}{conj.}{as, same way as}{*teng}{the same as}{*fya}{path}{S}
\cww{tengkrr}{tEN\uline{k\textsyllabic{r}}}{conj.}{while, same time as}{krr}{time}{S}
\cww{teng'u}{tENPu}{pn.}{same thing}{'u}{thing}{?}
\word{teri--}{tE\uline{ri}}{adp.}{about, concerning}{FE}
\word{terkup}{\uline{t.ER}k.up\textcorner}{v.}{die}{S}
\cww{teswot\`ing}{tE\uline{swo}t.IN}{v.}{grant}{t\`ing}{give}{S}
\word{teya}{tEja}{adj.}{full}{S}
\word{teylu}{tEjlu}{n.}{beetle larva (Pandoran fauna)}{C,S}
\word{tewti}{tEwti}{intj.}{wow}{FE}
\affix{--ti}{ti}{accusative}{pronouns and nouns}{W}
\word{til}{til}{n}{joint, hinge}{S}
\word{Tipnai}{tipnai}{prop.n.}{a Na'vi clan name}{G}
\word{tirea}{ti\uline{RE}a}{n.}{spirit}{S}
\cw{tireafya'o}{tiR\uline{E}a\uline{fja}Po}{n.}{spirit path}{tirea}{spirit}{fya'o}{path}{S}
\cw{tireaioang}{tiR\uline{E}ai\uline{o}aN}{n.}{spirit animal}{tirea}{spirit}{ioang}{animal}{S}
\cw{tireap\"angkxo}{tiREap.\ae Nk'.o}{v.}{commune with}{tirea}{spirit}{p\"angkxo}{chat}{F}
\derivingaffix{t\`i--}{tI}{noun}{W}
\word{t\`i'eyng}{tIPEjN}{n.}{answer}{F}
\word{t\`ifmetok}{tIf\uline{mE}tok\textcorner}{n.}{test}{S, MS}
\word{t\`iftang}{tIftaN}{n.}{stopping}{S}
\word{t\`ifyaw\`intxu}{tIfjawInt'u}{n.}{guidance}{F}
\word{t\`ihawnu}{tI\uline{haw}nu}{n.}{protection}{S}
\word{t\`ikangkem}{tIkaNkEm}{n.}{work}{F}
\word{t\`ikawng}{tI\uline{kawN}}{n.}{evil}{S,MS}
\word{t\`ikenong}{tI\uline{kE}noN}{n.}{example}{S}
\word{t\`ikin}{tIkin}{n.}{need}{S}
\derives{t\`ilor}{tIloR}{n.}{beauty}{somt\`ilor}{hot beauty (Pandoran flora)}{SG}
\word{t\`ing}{t.IN}{v.}{give}{S}
\word{t\`ing mikyun}{t.Im \uline{mik}jun}{v.}{listen, lit. give an ear}{S}
\word{t\`ing nari}{t.In \uline{na}Ri}{v.}{look, lit. give an eye}{S}
\word{t\`ingay}{tINaj}{n.}{truth}{S}
\cww{t\`ipawm}{tIpawm}{n.}{question}{pawm}{ask}{FE}
\word{t\`iran}{t.IR.an}{v.}{walk}{S}
\cww{t\`iyawn}{tIjawn}{n.}{love}{yawne}{beloved}{F}
\word{t\`irey}{tIREj}{n.}{life}{S}
\word{t\`irol}{tIRol}{n.}{song}{S}
\cww{t\`ispxin}{tIsp'in}{n.}{the state of being ill}{spxin}{sick}{FE}
\word{t\`itakuk}{tItakuk\textcorner}{n.}{strike}{G}
\word{t\`itxur}{tIt'uR}{n.}{strength}{S}
\marker{to}{to}{conj. {\it than}, comparative marker}{S}
\derive{*to}{to}{prep.}{away from}{neto}{away (direction)}{m\`iso}{away (direction)}{T}
\word{tok}{t.ok\textcorner}{v.}{be at, occupy, exist}{S}
\loan{toktor}{\uline{tok}toR}{n.}{doctor}{doctor}{S}
\word{tokx}{\uline{tok'}}{n.}{body}{S, MS}
\word{tompa}{tompa}{n.}{rain}{S}
\word{toruk}{\uline{to}Ruk\textcorner}{n.}{great leonopteryx, last shadow}{C,S}
\cw{toruk makto}{toRuk makto}{n.}{toruk rider}{toruk}{leonopteryx}{makto}{ride}{C}
\word{torukspxam}{toRuksp'am}{n.}{octoshroom, great leonopteryx fungus (Pandoran flora)}{SG}
\word{trr}{t\textsyllabic{r}}{n.}{day}{S}
\word{trram}{t\textsyllabic{r}am}{adv.}{yesterday}{F}
\derivingaffix{--tu}{tu}{direct object noun}{}, as in {\bf spe'etu} {\it captive} $_T$ %fix
\word{*tu}{tu}{n.}{person, personage marker}{T}
\word{tukru}{tuk\uline{Ru}}{n.}{spear}{S}
\word{tul}{t.ul}{v.}{run}{S}
\word{tung}{t.uN}{v.}{allow}{S}
\lenite{tupe}{\uline{tu}pE}{inter.}{who}{pesu}{S}
\word{tut}{tut\textcorner}{part.}{continuation marker}{FE}
\word{tute}{\uline{tu}tE}{n.}{person}{F}
\word{tute}{tut\uline{E}}{n.}{female}{F}
\word{tutean}{tutE\uline{an}}{n.}{male}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{Ts}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{tsa}{\texttslig a}{dem.}{that}{S}
\derive{*tsa--}{\texttslig a}{adp.}{that}{tsakem}{that action}{tsakrr}{that time}{T}
\cw{tsa'u}{\texttslig aPu}{dem.}{that (thing)}{tsa}{that}{*'u}{thing}{S} $_T$
\word{tsahaylu}{\texttslig a\uline{hEj}lu}{n.}{bond}{C,S} (changed pronunciation for phonetic reasons) $_T$
\word{Tsah\`ik}{\uline{\texttslig a}hIk\textcorner}{prop.n.}{matriarch, high priestess, interpreter}{S}
\word{tsakem}{\texttslig akEm}{dem.}{that action}{M}
\word{tsakrr}{\texttslig ak\textsyllabic{r}}{dem.}{that time}{F}
\word{tsam}{\texttslig am}{n.}{war}{S}
\cw{tsampongu}{\uline{\texttslig am}poNu}{n.}{war party}{tsam}{war}{pongu}{party}{S}
\cw{tsamsiyu}{\uline{\texttslig am}siju}{n.}{warrior}{tsam si}{make war}{--yu}{agent suffix}{S}
\word{tsap'alute}{\texttslig apP\uline{a}lutE}{n.}{apology}{S}
\cw{tsatseng}{\texttslig a\texttslig EN}{dem.}{there, that place}{tsa}{that}{tseng}{place}{S}
\cw{tsatu}{\texttslig atu}{dem.}{that person}{tsa}{that}{tu}{person}{S}
\word{tsawke}{\texttslig awkE}{n.}{sun}{S}
\word{tsawl}{\texttslig awl}{adj.}{big (in stature), tall}{S}
\word{tsawlapxangrr}{\texttslig awlap'aN\textsyllabic{r}}{n.}{unidelta tree, lit. tall large root (Pandoran flora)}{SG} (for short, {\bf tsawlapx})
\derive{*tsay+}{\texttslig aj}{adp.}{those}{*tsa--}{that}{ay+}{plural}{D}
\word{tse'a}{\texttslig.EP.a}{v.}{see (physical sense)}{S}
\word{tseng(e)}{\texttslig EN(E)}{n.}{place}{S}
\lenite{tsengpe}{\texttslig ENpE}{inter.}{where}{peseng}{S}
\word{tseo}{\texttslig Eo}{n.}{art}{S}
\word{ts\`ing}{\texttslig IN}{num.}{four}{S}
\word{Tskaha}{\texttslig kaha}{n.}{tribe name}{C, M}
\word{tskalep}{\texttslig kalEp\textcorner}{n.}{crossbow}{G}
\word{tsko}{\texttslig ko}{n.}{bow (weapon)}{S}
\word{tsko swizaw}{\texttslig ko swizaw}{n.}{bow and arrows}{S}
\word{tskxe}{\texttslig k'E}{n.}{rock, stone}{S}
\word{tskxekeng}{\texttslig k'EkEN}{n.}{training, exercise}{S}
\word{tslam}{\texttslig l.am}{v.}{understand}{S}
\word{tsleng}{\texttslig lEN}{adj.}{false}{S}
\derives{tslolam}{\texttslig lolam}{intj.}{got it, I understand}{tsl$<$ol$>$am}{understand}{S} with perfective aspect infix $_T$
\word{tsmuk}{\texttslig muk\textcorner}{n.}{sibling}{S}
\word{tsmuktu}{\uline{\texttslig muk}tu}{n.}{sibling}{S} (allomorph of {\bf tsmuk} {\it sibling}) $_T$
\word{tsmukan}{\uline{\texttslig muk}an}{n.}{brother}{S,MS}
\word{tsmuke}{\texttslig mukE}{n.}{sister}{S}
\word{tsn\`i}{\texttslig nI}{}{that (function word)}{S}
\derives{*tsngal}{\texttslig Nal}{n.}{cup}{tawtsngal}{sky cup}{T}
\word{tspang}{\texttslig p.aN}{v.}{kill}{S}
\word{tsranten}{\texttslig R.ant.En}{v.}{matter, be of importance}{S}
\word{tsteu}{\texttslig tEu}{adj.}{brave}{S} (might be {\bf tstew})$_D$
\word{tstxo}{\texttslig t'o}{n.}{name}{FE}
\word{tsun}{\texttslig.un}{v.}{can, be able}{S}
\word{Tsu'tey}{\texttslig uP\uline{tEj}}{prop.n.}{man's name}{C}
\word{tswon}{\texttslig w.on}{v.}{fly}{S} (Corrected from {\bf tsw$<$ay$>$on}) $_T$
\word{tsyal}{\texttslig jal}{n.}{wing}{S}
\derives{*tsyo}{\texttslig jo}{n.}{flour}{tsyorina'wll}{flour seed plant}{T}
\cw{tsyorina'wll}{\texttslig joRinaPw\textsyllabic{l}}{n.}{cycad, flour seed plant (Pandoran flora)}{rina'}{seed}{*wll}{plant}{SG} $_T$
\end{hangparas} }\end{multicols}
\noindent\textbf{Tx}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{txan}{t'an}{adj.}{great, much, many}{S}
\word{txantsan}{\uline{t'an}tsan}{adj.}{excellent}{FE}
\cw{txantslusam}{\uline{t'an}\texttslig lusam}{adj.}{wise}{txan}{much}{tsl$<$us$>$am}{understand}{S}
\word{txe'lan}{t'EP\uline{lan}}{n.}{heart}{S,MS}
\word{txele}{t'ElE}{n.}{matter (subject)}{S}
\word{txen}{t'En}{adj.}{awake}{S}
\word{txep}{t'Ep\textcorner}{n.}{fire}{S}
\word{txewk}{t'Ewk\textcorner}{n.}{club (weapon)}{G}
\word{txey}{t'.Ej}{v.}{halt, stop}{S}
\word{tx\`im}{t'Im}{n.}{butt, rear end}{S}
\word{tx\`ing}{t'.IN}{v.}{leave, abandon}{S}
\word{txll'u}{t'\textsyllabic{l}Pu}{n.}{hookagourd (Pandoran flora)}{SG}
\word{txo}{t'o}{conj.}{if}{S}
\word{txoa}{t'oa}{n.}{forgiveness}{S}
\cww{txokefyaw}{\uline{t'ok}Efjaw}{conj.}{if not, or else}{txo-ke-fya-aw}{if not one path}{S}
\word{txon}{t'on}{n.}{night}{S}
\word{txopu}{t'opu}{}{fear}{S}
\word{txula}{\uline{t'.u}l.a}{v.}{build}{FE}
\word{txum}{t'um}{n.}{poison}{S}
\cw{txumpaywll}{t'umpayw\textsyllabic{l}}{n.}{scorpion thistle, poison water plant (Pandoran flora)}{txum}{poison}{paywll}{water plant}{SG}
\word{txumts\"a'wll}{t'um\texttslig\ae Pw\textsyllabic{l}}{n.}{reclining flask (Pandoran flora)}{SG}
\word{txur}{t'uR}{adj.}{strong}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{U}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{u'imi}{uPimi}{n.}{streamers}{SG}
\word{ulte}{\uline{u}ltE}{conj.}{and (connects two clauses)}{S}
\word{ultxa}{ult'a}{n.}{meeting}{S}
\word{unil}{\uline{u}nil}{n.}{dream}{S}
\word{Uniltaron}{\uline{u}nil\uline{ta}Ron}{prop.n.}{Dream Hunt}{C,S,MS}
\word{unilt\`irantokx}{uniltI\uline{Ran}tok'}{n.}{avatar, dreamwalker body}{S}
\word{unilt\`iranyu}{unitI\uline{Ran}ju}{n.}{dreamwalker}{S}
\alloffix{--ur}{uR}{--ru}{dative}{consonant}{W}
\infix{$<$us$>$}{us}{participle}{1}{W}
\derive{*--ut}{ut\textcorner}{n.}{tree}{pxiut}{sharp tree}{rumut}{ball tree}{SG} (compound-word allomorph of {\bf utral}) $_T$
\word{utral}{utRal}{n.}{tree}{S}
\word{Utral Aymokriy\"a}{\uline{u}tRal aj\uline{mok}Rij\ae}{prop.n.}{Tree of Voices (Pandoran flora)}{S} (from {\bf utraya mokri} $_C$)
\word{utral utu mauti}{utRal utu mauti}{n.}{push fruit tree (Pandoran flora)}{SG}
\word{utu mauti}{utu mauti}{n.}{banana fruit, push fruit (Pandoran flora)}{SG}
\word{uvan}{uvan}{n.}{game}{FE}
\word{uvan si}{uvan s.i}{v.}{play}{FE}
\infix{$<$uy$>$}{uj}{honorific or ceremonial}{3}{F}
\end{hangparas} }\end{multicols}
\noindent\textbf{V}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{vay--}{vaj}{adp.}{up to}{F}
\word{vir\"a}{v.iR.\ae}{v.}{spread, proliferate}{S} (possible loan from English {\it virus}) $_T$
\derive{*vitra}{vitRa}{n.}{soul}{Vitrautral}{Tree of Souls}{Vitraya Ramunong}{Well of Souls}{D}
\cww{Vitrautral}{vitRautRal}{n.}{Tree of Souls}{utral}{tree}{F}
\word{Vitraya Ramunong}{vitRaja RamunoN}{prop.n}{Well of Souls}{C}
\word{vol}{vol}{num.}{eight}{FE}
\word{vrrtep}{\uline{v\textsyllabic{r}}tEp}{n.}{demon}{S}
\word{vul}{vul}{n.}{branch (of a tree)}{S}
\end{hangparas} }\end{multicols}

\noindent\textbf{W}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{way}{waj}{n.}{song}{S}
\word{waytelem}{\uline{waj}tElEm}{n.}{song chord}{S}
\word{win}{win}{adj.}{fast}{SG}
\word{w\`intxu}{w.Int'.u}{v.}{show}{S}
\derive{*wll}{w\textsyllabic{l}}{n.}{plant}{pamtseowll}{music plant}{txumts\"a'wll}{poison-squirting plant}{T}
\derive{*wong}{woN}{n.}{sameness, the same}{kewong}{alien}{ketuwong}{alien}{T}
\word{wrrpa}{w\textsyllabic{r}pa}{adj.}{outside}{S}
\word{wutso}{wu\texttslig o}{n.}{dinner, served meal}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{Y}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{ya}{ja}{n.}{air}{FE}
\alloffix{--y\"a}{j\ae}{\"a, ey\"a}{genitive}{vowel}{W}
\word{yawne}{jawnE}{adj.}{beloved}{F}
\word{ye'r\`in}{\uline{jE}PRIn}{adv.}{soon}{S,MS}
\derives{*ye'ung}{jePuN}{n.}{sanity}{keye'ung}{insanity}{D}
\word{yerik}{\uline{jE}Rik\textcorner}{n.}{hexapede (Pandoran fauna)}{S}
\word{yey}{jEj}{adj.}{straight}{S}
\word{y\`im}{j.Im}{v.}{bind}{S}
\word{yol}{jol}{adj.}{long (period of time)}{S}
\word{yom}{j.om}{v.}{eat}{S}
\derive{yomh\`i'ang}{jomhIPaN}{n.}{small animal/insect eater, dakteron (Pandoran flora)}{h\`i'i}{small}{ioang}{animal}{SG} $_T$
\cw{yomioang}{jomi\uline{o}aN}{n.}{chalice plant (Pandoran flora)}{yom}{eat}{iaong}{animal $_T$}{SG}
\cw{yomt\`ing}{\uline{jom}t.IN}{v.}{feed}{yom}{eat}{t\`ing}{give}{S}
\derivingaffix{--yu}{ju}{nominative agent noun}{W}, as in {\bf taron-yu} {\it hunter}
\word{yur}{j.uR}{v.}{wash}{S}
\end{hangparas} }\end{multicols}
\noindent\textbf{Z}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\derive{*za}{z.a}{v.}{go, movement}{za'u}{come}{za'\"ar\`ip}{pull}{T}
\word{za'\"ar\`ip}{zaP.\uline{\ae}R.Ip\textcorner}{v.}{pull}{S}\
\word{za'u}{\uline{z.a}P.u}{v.}{come}{S}
\word{zam}{zam}{num.}{sixty four}{F}
\derive{zamunge}{za\uline{m.u}N.E}{v.}{bring}{*za}{go}{munge}{bring}{S}
\word{zawng}{zawN}{n.}{shriek}{G}
\word{zekw\"a}{zEkw\ae}{n.}{finger}{S}
\word{zene}{z.En.E}{v.}{must}{S}
\word{z\`is\`it}{\uline{zI}sIt\textcorner}{n.}{year}{S}
\word{zong}{z.oN}{v.}{save}{S}
\cw{zongtseng}{zoN\texttslig EN}{n.}{safe place, refuge}{zong}{save}{tseng}{place}{S}
\word{zoplo}{zop\uline{lo}}{n.}{offence, insult}{S}
\cw{zusawkrr}{zusawk\textsyllabic{r}}{n.}{future, in the future}{z$<$us$>$a'u}{coming}{krr}{time}{FE}
\end{hangparas} }\end{multicols}
\newpage
\appendix
\section{The Number System.}
The following are charts of Na'vi numbers. Note that it is not a decimal, but an octal system. Underlining indicates where the stress ought to fall. 
\begin{center}
\ \\
\begin{supertabular}{|c|c|c|c|c|c|c|} 
\hline
\multicolumn{6}{|c|}{Na'vi Numbers: 0 -- 39} \\ \hline
 & 0 & 1 & 2 & 3	 & 4\\ \hline
0 & & vol & \uline{me}vol & \uline{pxe}vol & \uline{ts\`i}vol	\\
1 & 'aw & vol\uline{aw} & mevol\uline{aw} & pxevol\uline{aw} & ts\`ivol\uline{aw} \\
2&\uline{mu}ne&vo\uline{mun}	&mevo\uline{mun}&pxevo\uline{mun} & ts\`ivo\uline{mun}\\
3&pxey&vo\uline{pey}&mevo\uline{pey}&pxevo\uline{pey} & ts\`ivo\uline{pey}	\\
4 & ts\`ing & vo\uline{s\`ing} & mevo\uline{s\`ing} & pxevo\uline{s\`ing} & ts\`ivo\uline{s\`ing}\\
5 & mrr & vo\uline{mrr} & mevo\uline{mrr} & pxevo\uline{mrr} & ts\`ivo\uline{mrr}\\
6 & \uline{pu}kap & vo\uline{fu} & mevo\uline{fu} & pxevo\uline{fu}&ts\`ivo\uline{fu}\\
7 & \uline{ki}n\"a & vo\uline{hin} & mevo\uline{hin} & pxevo\uline{hin} & ts\`ivo\uline{hin}\\
\hline 
\end{supertabular} \end{center} 
\begin{center}
\begin{supertabular}{|c|c|c|c|} 
\hline
\multicolumn{4}{|c|}{Na'vi Numbers: 40 -- 63} \\ \hline
 & 5 & 6 & 7\\ \hline
0 & \uline{mrr}vol & \uline{pu}vol & \uline{ki}vol	\\
1 & mrrvo\uline{law} & puvo\uline{law} & kivo\uline{law}\\
2 & mrrvo\uline{mun} & puvo\uline{mun} & kivo\uline{mun}\\
3 & mrrvo\uline{pey} & puvo\uline{pey} & kivo\uline{pey}\\
4 & mrrvo\uline{s\`ing} & puvo\uline{s\`ing} & kivo\uline{s\`ing}\\
5 & mrrvo\uline{mrr} & puvo\uline{mrr} & kivo\uline{mrr}\\
6 & mrrvo\uline{fu} & puvo\uline{fu} & kivo\uline{fu}\\
7 & mrrvo\uline{hin} & puvo\uline{hin} & kivo\uline{hin}\\
\hline 
\end{supertabular} \end{center}
\begin{center}
\ \\ \ \\
\begin{supertabular}{|c|c|c|c|c|c|c|} 
\hline
\multicolumn{7}{|c|}{More Na'vi Numbers} \\ \hline
 & x1 & combining & x8 & x64 & x512 & x4096\\ \hline
1 & 'aw & -aw & vol & zam & \uline{vo}zam & \uline{za}zam\\
2 & \uline{mu}ne & -mun & \uline{me}vol & \uline{me}zam & \uline{me}vozam & \uline{me}zazam\\
3 & pxey & -pey & \uline{pxe}vol & \uline{pxe}zam & \uline{pxe}vozam & \uline{pxe}zazam\\
4 & ts\`ing&-s\`ing & \uline{ts\`i}vol & \uline{ts\`i}zam & \uline{ts\`i}vozam & \uline{ts\`i}zazam\\
5 & mrr & -mrr & \uline{mrr}vol & \uline{mrr}zam & \uline{mrr}vozam & \uline{mrr}zazam\\
6 & \uline{pu}kap & -fu & \uline{pu}vol & \uline{pu}zam & \uline{pu}vozam & \uline{pu}zazam\\
7 & \uline{ki}n\"a & -hin & \uline{ki}vol & \uline{ki}zam & \uline{ki}vozam & \uline{ki}zazam\\
\hline 
\end{supertabular} \end{center}
\newpage
\renewcommand{\infix}[7]{\par\textbf{#1}: [\textipa{#2}] $_{#5}$ #3 verbal infix in position #4: \textbf{#6} \textit{#7}}
\renewcommand{\affect}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3 affect verbal infix in position 3: \textbf{#5} \textit{#6}}
\renewcommand{\infixcw}[9]{\par\textbf{#1}: [\textipa{#2}] $_{#7}$ #3 verbal infix in position #4, derived from \textbf{#5} + \textbf{#6}: \textbf{#8} \textit{#9}}
\newcommand{\suffix}[7]{\par\textbf{#1}: [\textipa{#2}] $_{#5}$ #3 suffix for #4: \textbf{#6} \textit{#7}}
\newcommand{\prefix}[7]{\par\textbf{#1}: [\textipa{#2}] $_{#5}$ #3 prefix for #4: \textbf{#6} \textit{#7}}
\renewcommand{\alloffix}[8]{\par\textbf{#1}: [\textipa{#2}] $_{#6}$ #4 suffix for nouns ending in a #5 (allomorph of \textbf{#3}): \textbf{#7} \textit{#8}}
\renewcommand{\alloffixx}[8]{\par\textbf{#1}: [\textipa{#2}] $_{#6}$ #4 suffix for #5 (allomorph of \textbf{#3}): \textbf{#7} \textit{#8}}
\renewcommand{\derivingaffix}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3 deriving affix: \textbf{#5} \textit{#6}}
\renewcommand{\marker}[6]{\par\textbf{#1}: [\textipa{#2}] $_{#4}$ #3: \textbf{#5} \textit{#6}}
\section{Illegal Words}
The following words, or their roots, have at least one of the illegal sounds, spellings, or digraphs: 
\begin{center}'h [\textglotstop h], b [b], c [k], ch [t\textesh], d [d], g [g], h' [h\textglotstop], ii [i:], j [\textipa{Z}], sh [\textesh], uu [u:] \end{center}
\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{chan'tu gor'ek nuuto}{tSanPtu gorPEk\textcorner nuuto}{n.}{stone jar}{SG} $\phi$
\derives{chey}{tSEj}{n.}{possesions rack}{P'ah s'ivil chey}{rack}{SG} $\phi$
\word{Eywa k'sey nivi'bri'sta}{Ejwa kPsEj niviPbriPsta}{}{hammock }{SG}, from {\it Eywa cradles everyone} $\phi$
\word{feru m'predu'k}{fEru mPpREduPk\textcorner}{n.}{basket $\phi$}{SG}
\word{iveh k'nivi s'dir}{ivEh kPnivi sPdir}{n.}{baby sling}{SG} $\phi$
\loan{jakesully}{ZEjk'su:\textsyllabic{l}i}{n.}{moron, toruk maktoyu}{Jake Sully}{M} $\phi$
\word{lenay'ga}{lEnajPga}{n.}{slinger, predatory animal (Pandoran fauna) $\phi$}{SG}
\word{meresh'ti cau'pla}{mERESPti kauPpla] or [meReshPti kauPpla}{n.}{banshee catcher, bola, lit. nothing to see $\phi$}{SG}
\word{m'resh'tuyu}{mPRESPtuju] or [mPREshPtuju}{n.}{shield}{SG} $\phi$
\word{na'hla buk'ne}{naPhla bukPnE}{n.}{direhorse lead, lit. face pull $\phi$}{SG} 
\word{nikt'chey}{niktP\texttslig hEj] or [niktPt\textesh Ej}{n.}{small bit of food, packed lunch}{SG} $\phi$
\derives{nivi}{nivi}{n.}{hammock (informal), us}{Eywa k'sey nivi'bri'sta}{Ewya cradles everyone $\phi$}{SG}
\word{P'ah s'ivil chey}{pPah sPivil tSEy}{n.}{personal belongings rack}{SG} $\phi$
\word{pxtxawng}{p't'awN}{n.}{douchebag $\phi$}{C}
\loan{shah-tell}{SahtEl}{n.}{shuttle}{shuttle}{SG} $\phi$ $_T$
\word{su'shiri t'acto sa}{suPshiri tPacto sa] or [suPshiri tPacto sa}{n.}{toy ikran}{SG} $\phi$
\word{sumin jiit'luy}{sumin ZiitPluy}{n.}{leaf plate}{SG} $\phi$
\word{t'riti so jahmka}{tPRiti so Zahmka}{n.}{pendulum drum}{SG} $\phi$
\word{U'imi huyuticaya}{uPimi hujutikaja}{prop.n.}{a Na'vi clan name $\phi$}{SG}
\word{ulivi mari'tsey mak'dini'to}{Ulivi mariP\texttslig Ej makPdiniPto}{}{loom, lit. branches of the tree look to each other for strength, or many branches together are stong $\phi$}{SG}
\word{ulu'tah inib'sey mulsi}{uluPtah inibPsEy mulsi}{n.}{leaf plate}{SG} $\phi$
\end{hangparas} }\end{multicols}
Several words below break the following syllable rules. A glottal stop cannot go between two consonants. There can be no consonant clusters that do not start with any of the phonemes [f \texttslig\ s]. Only one consonant syllable finally.
\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\word{ar'lek}{arPlEk\textcorner}{n.}{a seed-bearing plant (Pandoran flora)}{G}
\word{Eywa s'ilivi mas'kit nivi}{Eywa sPilivi masPkit\textcorner nivi}{}{Ewya's wisdom is revealed to all of us}{SG}
\word{Eywa te' $\_$ tan'sey mak'ta}{Ejwa tEP $\_$ tanPsEJ makPta}{n.}{ikran queue harness, lit. the love of Eywa's embrace is gifted to (personal name)}{SG} (dubious: misspelling of te) $_T$
\word{k'sey nivi}{kPsEj nivi}{n.}{group sleep}{SG}
\word{Kyuna}{kjuna}{prop.n}{girl's name}{G}
\word{malmoks}{malmoks}{n.}{rings of stone}{G}
\word{mas'kit nivi sa'nok}{masPkit\textcorner nivi saPnok\textcorner}{}{mother loom}{SG} (dubious) $_T$
\word{mreki u'lito}{mREki uPlito}{n.}{fire pit}{SG}
\word{omati s'ampta}{omati sPampta}{n.}{blue flute}{SG}
\word{sumin'sey hulleh}{suminPsEj hu\textsyllabic{l}Eh}{}{sleeping outside of the group}{SG}
\word{tmi nat'sey}{tmi natPsEj}{n.}{bladder lantern}{SG}, from {\it food here}
\end{hangparas} }\end{multicols}
\newpage
\section{Inflections}
\noindent\textbf{Verb Infixes}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\infix{$<$am$>$}{am}{past tense}{2}{W}{Oe tamaron.}{I hunted.}
\infixcw{$<$arm$>$}{aRm}{imperfective past tense}{2}{am}{er}{W}{Oe tarmaron.}{I was hunting.}
\infix{$<$ay$>$}{aj}{future tense}{2}{W}{Oe tayaron.}{I will hunt.}
\affect{$<$\"ang$>$}{\ae N}{pejorative}{W}{Oe tar\"angon.}{I hunt (ugh!).}
\infix{$<$\"ap$>$}{\ae p}{reflexive}{1}{FE}{Oe t\"aparon.}{I hunt myself.}
\affect{$<$ei$>$}{Ei}{laudative}{F}{Oe tareion.}{I hunt (hooray!).}
\infix{$<$er$>$}{ER}{imperfective aspect}{2}{D}{Oe teraron.}{I am hunting.}
\infix{$<$ilv$>$}{ilv}{present perfective subjunctive}{2}{FE}{Oe new tslilvam.}{I want to understand.}
\infix{$<$imv$>$}{imv}{past subjunctive}{2}{FE}{Oe namew tslimvam}{I wanted to understand.}
\infix{$<$irv$>$}{iRv}{present imperfective subjunctive}{2}{FE}{Oe new tslirvam.}{I want to be understanding.}
\infix{$<$iv$>$}{iv}{subjunctive mood}{2}{W}{Ngal new futa oe tivaron.}{You want this thing: for me to hunt}
\infix{$<$iyev$>$}{ijEv}{future subjunctive}{2}{FE}{Oel kiyevame ngati.}{I will see you soon.}
\infix{$<$\`im$>$}{Im}{recent past}{2}{W}{Oe t\`imaron.}{I just hunted.}
\infix{$<$\`irm$>$}{IRm}{imperfective recent past tense}{2}{W}{Oe t\`irmaron.}{I was just hunting.}
\infix{$<$\`iy$>$}{Ij}{immediate future tense}{2}{W}{Oe t\`iyaron.}{I am about to hunt.}
\infix{$<$\`iyev$>$}{IjEv}{future subjunctive}{2}{FE}{Oel k\`iyevame ngati.}{I will see you soon.}
\infix{$<$ol$>$}{ol}{perfective aspect}{2}{W}{Oe tolaron.}{I hunted (finished action).}
\infix{$<$us$>$}{us}{participle}{1}{W}{Oe tusaron-a tute lu.}{I am a hunting person.} (Not sure) $_T$ 
\infix{$<$uy$>$}{uj}{honorific or ceremonial}{3}{F}{Na'viy\"a, l$<$uy$>$u hapx\`i.}{You are part of The People.}
\end{hangparas} }\end{multicols}
\noindent\textbf{Noun Inflections}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\alloffix{--\"a}{\ae}{ey\"a, y\"a}{genitive}{consonant}{W}{Oel yom ikran\uline{--\"a} yerikit.}{I eat the ikran's yerik.}
\suffix{--an}{an}{masculine}{certain nouns}{W}{Oe tute\uline{--an} lu.}{I am a male dude.}
\prefix{ay+}{aj}{plural}{nouns}{F}{\uline{Ay--}oe taron.}{We hunt.}
\suffix{--e}{E}{feminine}{certain nouns}{F}{Oe tut\uline{--e} ke lu.}{I am not a girl}
\alloffixx{=ey\"a}{Ej\ae}{\"a, y\"a}{genitive}{pronouns}{W}{Oel yom o\uline{ey\"a} yerikit.}{I eat my yerik}
\alloffix{--\`il}{Il}{l}{ergative case}{consonant}{W}{Ikran\uline{--\`il} taron yerikit.}{The ikran hunts the yerik.}
\alloffix{--\`iri}{IRi}{ri}{topical}{consonant}{W}{Ikran\uline{--\`iri} taron yerikit.}{The \uline{ikran} hunts a yerik.}
\alloffix{--it}{it\textcorner}{t, ti}{accusative}{consonant}{W}{Oel taron yerik\uline{--it}.}{I hunt the yerik.}
\suffix{--l}{l}{ergative case}{nouns ending in a vowel}{W}{Oe\uline{--l} taron yerikit.}{I hunt the yerik.}
\marker{ma}{ma}{part. vocative marker}{M}{\uline{Ma} tsmukan, kaltx\`i!}{Yo brother, hello!}
\prefix{me+}{mE}{dual}{nouns}{W}{Oel new \uline{me+}narit.}{I want eyes.}
\suffix{--ng}{N}{clusivity}{pronouns}{W}{Ayoe\uline{--ng} taron.}{We all hunt.}
\marker{pe+}{pE}{affix inter. marker \textit{what}}{W}{\uline{pe+}hrr nga tayaron.}{What time will you hunt?}
\prefix{pxe+}{p'E}{trial}{nouns and pronouns}{F}{Ayoeng \uline{Pxe}veng lu.}{We are three children.}
\suffix{--ri}{Ri}{topical}{nouns ending with a vowel}{F}{Oe\uline{--ri} taron yerikit.}{\uline{I} hunt the yerik.}
\alloffix{--ru}{Ru}{--ur}{dative}{vowel}{W}{Oel kxenerit nga\uline{--ru} t\`ing.}{I give fruit to you.}
\alloffix{--t}{t\textcorner}{it, ti}{accusative}{vowel}{W}{Oe\uline{--l} t\`irey\uline{--t} tslam.}{I understand life.}
\suffix{--ti}{ti}{accusative}{pronouns and nouns}{W}{Nga\uline{--ti} taron toruk\`il.}{The toruk hunts you.}
\alloffix{--ur}{ur}{--ru}{dative}{consonant}{W}{Ayeylanur oey\"a.}{To my friends.}
\alloffix{--y\"a}{j\ae}{\"a, ey\"a}{genitive}{vowel}{W}{Oel yom Neytiri\uline{-y\"a} yerikit.}{I eat Neytiri's yerik.}
\end{hangparas} }\end{multicols}
\noindent\textbf{Other}\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\marker{--a--}{a}{adjectival attributive marker}{F}{Oel ikranit \uline{a--}ean tswon.}{I fly the blue ikran.}
\derivingaffix{le--}{lE}{adjective}{F}{Poan \uline{le--}hrapp lu.}{He is dangerous.}
\derivingaffix{n\`i--}{nI}{adverbial}{W}{Oe \uline{n\`i--}Na'vi plltxe.}{I speak na'vi-ly. (Like a Na'vi.)}
\derivingaffix{*s\"a--}{s\ae}{instrumental noun}{T}{Oel ay\uline{s\"a--}numet nume.}{I learn the teachings.}
\derivingaffix{si}{si}{postpositional verb}{W}{Oe kelku \uline{si} m\`i kelutral.}{I live in Hometree. (lit. I home make in Hometree.)}
\derivingaffix{t\`i--}{tI}{noun}{W}{\uline{T\`i--}ngay s\`iltsan lu.}{The truth is good.}
\derivingaffix{--tu}{tu}{direct object noun}{T}{Oe taron\uline{--tu} lu.}{I am prey.}
\derivingaffix{--yu}{ju}{nominative agent noun}{W}{Oe ska'a\uline{--yu} ke lu.}{I am not a destroyer.}
\end{hangparas} }\end{multicols}
\newpage
\renewcommand{\infix}[5]{\par\textbf{#2}: #1. $_{#5}$ #3 verbal infix in position #4.}
\renewcommand{\affect}[4]{\par\textbf{#2}: #1. $_{#4}$ #3 affect verbal infix in position 3.}
\renewcommand{\infixcw}[5]{\par\textbf{#2}: #1. $_{#5}$ #3 verbal infix in position #4.}
\renewcommand{\affix}[5]{\par\textbf{#2}: #1. $_{#5}$ #3 affix for #4.}
\renewcommand{\alloffix}[6]{\par\textbf{#2}: #1. $_{#6}$ #4 suffix for nouns ending in a #5 (allomorph of \textbf{#3})}
\renewcommand{\derivingaffix}[4]{\par\textbf{#2}: #1. $_{#4}$ #3 deriving affix.}
\renewcommand{\marker}[4]{\par\textbf{#2}: #1. $_{#4}$ #3.}
\section{English Short-Hand Terms for Trilinear Glosses}
Here is a list of the short-hand notation often used in trilinear glosses on the Forums, with some common morphemes. 
\begin{multicols}{2}{\begin{hangparas}{.5cm}{1}\noindent
\alloffix{--t}{ACC}{ti}{accusative}{vowel}{W}
\affix{--ti}{ACC}{accusative}{nouns}{W}
\alloffix{--it}{ACC}{ti}{accusative}{consonant}{W}
\marker{--a--}{ADJ M.}{adjectival attributive marker}{F}
\derivingaffix{le--}{ADJ}{adjective}{F}
\derivingaffix{n\`i--}{ADV}{adverbial}{W}
\derivingaffix{--yu}{AGENT N}{nominative agent noun}{W}
\affix{--ng}{CLU}{clusivity}{pronouns}{W}
\affix{--ru}{DAT}{dative}{nouns}{W}
\affix{me+}{DUAL}{dual}{nouns}{W}
\alloffix{--\`il}{ERG}{--l}{ergative case}{consonant}{W}
\affix{--l}{ERG}{ergative case}{nouns ending in a vowel}{W}
\marker{--e}{FEM}{feminine suffix for certain nouns}{F}
\infix{$<$ay$>$}{FUT}{future tense}{2}{W}
\alloffix{--\"a}{GEN}{y\"a}{genetive}{consonant}{W}
\alloffix{--y\"a}{GEN}{\"a}{genitive}{vowel}{W}
\infix{$<$\`iy$>$}{IMM. FUT}{immediate future tense}{2}{W}
\infix{$<$er$>$}{IMPF}{imperfective aspect}{2}{D}
\infixcw{$<$arm$>$}{IMPF. PAST}{imperfective past tense}{2}{W}
\infixcw{$<$\`irm$>$}{IMPF. REC. PAST}{imperfective recent past tense}{2}{W}
\derivingaffix{*s\"a--}{INST}{instrumental noun}{T}
\marker{pe+}{INTER}{affix inter. marker \textit{what}}{W}
\affect{$<$ei$>$}{LAUD}{laudative}{F}
\marker{--an}{MASC}{masculine suffix for certain nouns}{W}
\derivingaffix{t\`i--}{NMLZ}{nominalizer (noun-er)}{W}
\derivingaffix{--tu}{OBJ}{direct object noun}{T}
\infix{$<$us$>$}{PART}{participle}{1}{W}
\infix{$<$am$>$}{PAST}{past tense}{2}{W}
\affect{$<$\"ang$>$}{PEJ}{pejorative}{W}
\infix{$<$ol$>$}{PERF}{perfective aspect}{2}{W}
\affix{ay+}{PLU}{plural}{nouns}{F}
\infix{$<$\`im$>$}{REC. PAST}{recent past}{2}{W}
\infix{$<$iv$>$}{SUBJ}{subjunctive mood}{2}{W}
\alloffix{--\`iri}{TOP}{ri}{topical}{consonant}{W}
\affix{--ri}{TOP}{topical}{nouns ending with a vowel}{F}
\derivingaffix{si}{V}{postpositional verb}{W}
\marker{ma}{VOC}{part. vocative marker}{F}
\end{hangparas} }\end{multicols}
\newpage
\section{The Change Log}
Subscript indicate who helpfully pointed out errors. Old logs versions are \href{http://forum.learnnavi.org/intermediate/my-dictionary/}{here}.
\begin{itemize}
\item 7.2 changed all pe+ or --pe words to POS inter.
\item 7.309 indicated place of input for all verb infixes.  Edited {\bf K\`iyevame, futa}, infix $<$iv$>$ example, {\bf fwa, k\`im, ngop, pe'un}, added {\bf *ye'ung} $_{Plumps83}$, changed {\bf tsakem}
\item 8 added Appendix A, the Number System. Also, {\bf zusawkrr} {\it future} $_{Nayumeie}$, added {\bf frrfen, frrtu, teri, letsunslu} {\it visit, guest, about, possible}, edited {\bf k\`iyevame}, added lots of infixes in the {\bf i} section, edited the infix examples. $_{suomichris}$, reclassified all {\bf f\`i-, tsa-} relevant words as demonstratives. 
\item 8.105 added Appendix D: English Trilinear Short-hand gloss terms, edited Number Graph issue $_{tikawnga}$, added {\bf eyawr, run} {\it correct, find/discover} $_{Mirri}$, changed all {\bf \`iyve} entries to {\bf iyev}, fixed number chart ' issue, $_{Lance}$, edited {\bf na, si} $_{Mirri}$, edited {\bf si} example
\item 8.23 Moved \`i and \"a into their own categories. :), edited {\bf mune}, added {\bf pxey, mrr, pukap, zam} {\it 3, 5, 6, 64}, fixed IPA errors in {\bf rel, reltseo, Rai'uk, t\`ilor, tireap\"angkxo, tsyorina'wll, utral uta mauti, vitra, vitraya ramunong, vitrautral, frakrr, *kar, Nera'nak, ohe, Anurai}, added stress for {\bf \uline{pxe+}, \uline{ki}n\"a, \uline{pu}kap}, \uline{me}+, to all of the number chart, fixed underlining ugliness $_{T}$, added {\bf lefpom} {\it happy, peaceful, joyous (not for people)} $_{roger}$, added {\bf kalin, ftx\`i, ya, prrnesyul, prrnen, syulang, spxin, s\"aspxin, t\`ispxin, mi, mungwrr--, uvan, uvan si} {\it sweet, tongue, air, bud, infant/baby, flower, sick, disease, the state of being sick, yet/still, except, game, to play}, edited {\bf s\"a--, fu} $_{Kiliya}$, alphabetized some {\bf \`i--, t\`i--, n\`i--} entries properly, added {\bf talun(a), tawkey(a), lun, oeyk, holpxay, hol, tafral} {\it because, because, reason, cause, number, few, therefore} $_{P}$, fixed IPA in {\bf iyev} entry and example. $_{Atantsawkeya}$
\item 9 - fixed unicode input for {\bf \"a} and {\bf \`i}. The document should now be searchable.
\item 9.103 - added {\bf reltseotu} {\it artist} $_{Lance}$, fixed {\bf -\`il, \"ang} examples. Added {\bf *ngrr} {\it root} $_{Utraya}$ $_{Mokri}$, edited {\bf vay--} $_{roger}$, {\bf futa}, deleted {\bf *'it} $_{suomichris}$, added {\bf $<$\"ap$>$, tstxo, tut} reflexivity infix, {\it name}, continuation particle $_{Tseyk}$, added an unreleased mark to every IPA word-final voiceless stop, with Tuiq's amazing coding help. changed fyapo to frapo in preamble $_{Plumps83}$, edited {\bf *ngrr, vol} $_{Utraya}$
\item 9.15 realphabetised li--, l\`i--, deleted vofu, ts\`ivol $_{mirri}$, fixed IPA for {\bf iyev, tireap\"angkxo, t\`itakuk, --ur}, fixed examples $_{okristi}$, fixed {\bf tengfya} $_{Alim}$, changed infix positions for {\bf ralp.eng} $_{plumps83}$, corrected IPA {\bf lok, mip, kip, kop, pak, swot, tut}, changed adp. into affix and {\bf ma} in english terms appendix, fixed suffix/affix marking for noun inflections appendix
\end{itemize}
\end{document}